---
title: "Benchmarking SaintExpress"
author: "Witold Wolski"
date: "12/2/2021"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      toc_collapsed: false
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Benchmarking SaintExpress} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}

---



```{r setup, include=FALSE}
evalAll <- TRUE
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Importing MQ data into proLFQua


First read data and annotation.

```{r fromPeptideTXT , eval = evalAll}
library(prolfqua)
datadir <- file.path(find.package("prolfquadata") , "quantdata")
inputMQfile <-  file.path(datadir,
                          "MAXQuant_IonStar2018_PXD003881.zip")
data <- tidyMQ_Peptides(inputMQfile)

inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")
annotation <- readxl::read_xlsx(inputAnnotation)
```

Create `proLFQua` configruation for MQ `peptide.txt` file and annotate data. 

```{r}
config <- prolfqua::create_config_MQ_peptide()
res <- add_annotation(
  data,
  annotation,
  fileName = "raw.file"
)

config$table$factors[["dilution."]] = "sample"
config$table$factors[["run_Id"]] = "run_ID"
config$table$factorDepth <- 1
data <- setup_analysis(res, config)
```

Setup LFQData class, filter, transform and aggregate peptide intensities:
- remove proteins with only one peptide identified (`filter_proteins_by_peptide_count`).
- remove small (zero) intensities (`remove_small_intensities`)
- log2 transform and scale intensities (`get_Transformer`)
- aggregate peptides to proteins (`get_Aggregator`)

```{r normalizeAndAggregate}

lfqdata <- prolfqua::LFQData$new(data,config)
lfqdata$data <- lfqdata$data |> dplyr::filter(!grepl("^REV__|^CON__", protein_Id)) 
lfqdata$filter_proteins_by_peptide_count()
lfqdata$hierarchy_counts()
lfqdata$remove_small_intensities()

tr <- lfqdata$get_Transformer()
subset_h <- lfqdata$get_copy()
subset_h$data <- subset_h$data |> dplyr::filter(grepl("HUMAN", protein_Id))
subset_h <- subset_h$get_Transformer()$log2()$lfq
lfqdataNormalized <- tr$log2()$robscale_subset(lfqsubset = subset_h)$lfq

lfqAggMedpol <- lfqdataNormalized$get_Aggregator()
lfqAggMedpol$medpolish()
lfqtrans <- lfqAggMedpol$lfq_agg
```

# Prepare protein data for SaintExpress analysis and run SaintExpress

SeintExpress requires specifying a control condition.  As control `C` we will use the samples with the lowest E-Coli spike in concentration (dilution `a`). Treatments `T` are samples with higher concentrations (dilution 'b', 'c', 'd', 'e').

- The code snipped below specifies control and treatment samples using `case_when`.
- adds protein lengths (`readPeptideFasta`,`addProteinLengths`).
- Creates _SaintExpress_ compatible inputs (`protein_2localSaint`).
- run saintExpress (`runSaint`)


```{r annotateProteins}

exampleDat <- lfqtrans$data |> dplyr::mutate(CorT = case_when(dilution. == "a" ~ "C", TRUE ~ "T"))
exampleDat$protein_Id <- gsub("~.*","", exampleDat$protein_Id)

# sample protein lengths


Ecolifasta <- system.file("fastaDBs/uniprot-proteome_UP000000625_reviewed_yes.fasta.gz",package = "prolfquadata")
Humanfasta <- system.file("fastaDBs/uniprot-proteome_UP000005640_reviewed_yes.fasta.gz",package = "prolfquadata")

Ecolifasta <- prozor::readPeptideFasta(Ecolifasta)
Humanfasta <- prozor::readPeptideFasta(Humanfasta)
fasta <- c(Ecolifasta, Humanfasta)
exampleDat <- prolfqua::addProteinLengths(exampleDat, fasta)



res <- protein_2localSaint(exampleDat,quantcolumn = "medpolish",
                           proteinID = "protein_Id",
                           proteinLength = "protein.length",
                           IP_name = "raw.file",
                           baitCol = "dilution.",
                           CorTCol = "CorT"
)
stopifnot(names(res) == c( "inter", "prey",  "bait"))
resSaint <- runSaint(res,filedir = tempdir())

names(resSaint)
ctr <- prolfqua::ContrastsSaintExpress$new(resSaint$list)

```



```{r , fig.cap="Scatter plot showing the dependence of the SaintScore on the log2FC."}
pl <- ctr$get_Plotter()
pl$score_plot()$SaintScore
pl$histogram()

```


## Benchmarking

We benchmark _SaintExpress_ using the _Ionstar_ dataset. We are using here the lowest ECOLI spike in concentration as control.

```{r setUPBenchmark, eval = evalAll}
bb <- ctr$get_contrasts()
bb <- mutate(bb , PEP = 1-SaintScore)
ttd <- prolfqua::ionstar_bench_preprocess( bb , idcol = "Prey" )
benchmark_SaintExpres <- make_benchmark(ttd$data,
                                   contrast = "Bait",
                                fcestimate = "log2FC",
                                toscale = NULL,
                                benchmark = list(
                                  list(sc = "log2FC", desc = TRUE),    
                                  list(sc = "SaintScore", desc = TRUE),
                                  list(sc = "BFDR", desc = FALSE)
                                ),  
                                model_description = "SaintExpress_medpolishInt",
                                model_name = "SaintExpress_medpolishInt",
                                FDRvsFDP = list(list(sc = "BFDR", desc = FALSE))
, hierarchy = c("Prey"), summarizeNA = "SaintScore"
)
colnames(ttd$data)
sum(benchmark_SaintExpres$smc$summary$Prey)
sumarry <- benchmark_SaintExpres$smc$summary
table_facade(sumarry, caption = "nr of proteins with 0, 1, 2, 3 missing contrasts.")

```


```{r prepBenchmarkforComparison, include= FALSE, eval = evalAll}
xdd <- ttd$data |> dplyr::rename(protein_Id = Prey ,
                                  contrast = Bait,
                                  estimate = log2FC,
                                  statistic = SaintScore,
                                  p.value = PEP,
                                  FDR = BFDR  
)

benchmark2_SaintExpress <- make_benchmark(xdd, model_description = "SaintExpress_medpolishInt", model_name = "proDA")
```

```{r seveBenchresults, eval=FALSE, include=FALSE}
saveRDS(benchmark_SaintExpres, file = "../inst/Benchresults/benchmark_medpolish_SaintExpress.RDS")

```

```{r rocCurve, fig.cap="ROC curves", eval = evalAll}
res <- benchmark_SaintExpres$pAUC_summaries()
knitr::kable(res$ftable$content,caption = res$ftable$caption)
res$barp
```

```{r pAUC02, fig.cap="plot ROC curves", eval = evalAll}
#res$ftable
benchmark_SaintExpres$plot_ROC(xlim = 0.2)
```

```{r fdrfdp, fig.cap = "plot FDR vs FDP",eval = evalAll}
benchmark_SaintExpres$plot_FDRvsFDP()
```

```{r fdptpr,eval = evalAll}
benchmark_SaintExpres$plot_FDPvsTPR()
```

# Conclusions

There are several problems with the results produced by _SaintExpress_:

- First, the fold changes reported, have no relation with the true fold change (see volcano plots)
- Secondly, the score with the highest performance (largest pAUC) is the $\log_2(FC)$. However, since the _SaintScore_ also model the observed variances we would expect to perform better.
- Third, the performance of _SaintExpress_  is significantly worse when compared with other algorithms, e.g. _limma_ or _PRORA_ implemented in prolfqua or _proDA_ (see other benchmark vignettes).


# Saint Express BFDR.

_SaintExpress_ infers several statistics.
- the _SaintScore_ - Saint Probability, vaguely - probability that the protein is a true interactor.
- the BFDR Bayesian FDR.



The following article [http://varianceexplained.org/r/bayesian_fdr_baseball/] describes how the BFDR can be derived from the posterior error probability (PEP).
The BFDR equals the cumulative mean of the PEP.

_SaintExpress_ does not report the PEP. But if the _SaintScore_ is the probability that a protein is an interactor then, $1 - SaintScore$ can be interpreted as posterior error probability (that it is not an interactor) and hence we can compute the BFDR:

```{r defFunction}
computeFDR <- function(mdata ){
    mdata <- dplyr::mutate(mdata, PEP = 1 - SaintScore)
    mdata <- mdata |> arrange(PEP)
    mdata <- mdata |> mutate( myFDR = cummean(PEP))
    return(mdata)
}

```

There are two options, either to determine the BFDR for all Baits (first code snipped) or for each Bait (second one with the `for` loop).

```{r BFDRvsMyFDRglobal, fig.cap="BFDR vs myFDR for all Baits."}
reslist <- resSaint$list

reslist <- computeFDR(reslist)
ggplot(reslist, aes(x = BFDR, y = myFDR)) + geom_point() + geom_abline(slope = 1,color = "red")
```


```{r BFDRvsMyFDRlocal, fig.cap = "BFDR vs myFDR for each Bait."}
ad <- list()
for(i in unique(reslist$Bait)){
    print(i)
    set <- filter(reslist, Bait == i)
    ad[[i]] <- computeFDR(set)
}
ad <- bind_rows(ad)

ggplot(ad, aes(x = BFDR, y = myFDR)) + geom_point() + facet_wrap(~Bait) + ggplot2::geom_abline(slope = 1,colour = 2)
```

The SE BFDR is quite similar to the _FDR_ we estimated from the _SaintScore_. It also seems that _SaintExpress_ computes the _BFDR_ for all Baits not for each.
