---
title: "Comparing msqrob2 proDA and prolfqua using the CPTAC dataset"
author: "Witold E Wolski"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      toc_collapsed: false
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Comparing msqrob2 proDA and prolfqua using the CPTAC dataset} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Introduction

The r-package `r BiocStyle::Biocpkg("msdata")` msdata provides a subset of the data of the 6th study of the Clinical Proteomic Technology Assessment for Cancer (CPTAC). In this experiment, the Sigma Universal Protein Standard mixture 1 (UPS1) containing $48$ different human proteins was spiked in a protein background of 60 ng/$\mu L$ Saccharomyces cerevisiae strain BY4741. Two different spike-in concentrations were used, 6A ($0.25$ fmol UPS1 proteins/$\mu L$) and 6B ($0.74$ fmol UPS1 proteins/$\mu L$). Three replicates are available for each concentration.


```{r cars}
library(tidyverse)
tmp <- msdata::quant(full.names = TRUE)
xx <- read.csv(tmp, sep = "\t")
peptides <- prolfqua::tidyMQ_Peptides(xx)

annotation <- data.frame(raw.file = unique(peptides$raw.file))
annotation <- annotation |> dplyr::mutate(group = gsub("^6","",raw.file)) |> mutate(group = gsub("_[0-9]$", "", group))

peptides <- dplyr::inner_join(annotation, peptides)


config <- prolfqua::create_config_MQ_peptide()
config$table$factors["group."] = "group"

peptides <- prolfqua::setup_analysis(peptides, config)
lfqpeptides <- prolfqua::LFQData$new(peptides, config)
lfqpeptides$filter_proteins_by_peptide_count()
lfqpeptides$remove_small_intensities()

lfqpeptides$hierarchy_counts()

lfqpeptides$factors()
tmp <- lfqpeptides$data |> filter(grepl("_UPS", protein_Id))

```

# Transform and Aggregate

First we transform the data using vsn::justvsn function and afterwards we estimate protein intensities using
Tukeys median polish.

```{r}
tr <- lfqpeptides$get_Transformer()
tr <- tr$intensity_matrix(vsn::justvsn)
lfqtransformed <- tr$lfq
agg <- lfqtransformed$get_Aggregator()
agg$medpolish()
pl <- agg$plot()
```

```{r aggregationResults, fig.cap = "Peptide abundances and protein abundance estimates."}
gridExtra::grid.arrange(grobs = pl$plots[33:38])
lfqProt <- agg$lfq_agg

```

# Fit model with prolfqua

```{r}
library(prolfqua)

model <- paste(lfqProt$response(), " ~ group. ")
lm <- prolfqua::strategy_lm(model)
lmmodl <- prolfqua::build_model(lfqProt, lm)
contrast <- c("b_vs_a" = "group.b - group.a")

clm <- prolfqua::Contrasts$new(lmmodl, contrast)
tt <- clm$get_contrasts()

csimp <- prolfqua::ContrastsSimpleImpute$new(lfqProt, contrast)
tt <- csimp$get_contrasts()

merge <- prolfqua::merge_contrasts_results(clm, csimp)
cmod <- ContrastsModerated$new(merge$merge)

pl <- cmod$get_Plotter()
pl$volcano()$FDR

```

Setting up benchmark. UPS proteins are true positives, while YEAST proteins are true negatives.
Contaminants are removed. This leaves 1460 proteins.

```{r}
ttd <- prolfquabenchmark::cptac_bench_preprocess(cmod$get_contrasts(), idcol = "protein_Id")
benchmark_merged <- prolfqua::make_benchmark(
  ttd$data,
  model_description = "merge of prot moderated and imputed",
  model_name = "prolfqua_merged")


sumarry <- benchmark_merged$smc$summary
knitr::kable(sumarry, caption = "nr of proteins with 0, 1, 2, 3 missing contrasts.",format = "html")

```

```{r}
res <- benchmark_merged$pAUC_summaries()
knitr::kable(res$ftable$content,caption = res$ftable$caption)
res$barp
```




# Fit model with proDA

```{r fitProDAmodel}
library(prolfqua)
library(proDA)
library(SummarizedExperiment)
se <- prolfqua::LFQDataToSummarizedExperiment(lfqProt)

mm <- model.matrix(~ group., colData(se))
proModel <- proDA::proDA(se, design = mm, data_is_log_transformed = TRUE)
res <- test_diff(proModel, "group.b", sort_by = "pval")
res$contrast = "b_vs_a"
tmp <- prolfqua::ContrastsProDA$new(res, c("a_vs_b" = "group.b - group.a"))
tmp$get_Plotter()$volcano()$adj_pval

```




```{r setupproDABenchmark}


ttd <- prolfquabenchmark::cptac_bench_preprocess(tmp$get_contrasts(), idcol = "name")
benchmark_proDA <- prolfqua::make_benchmark(ttd$data,
                                            contrast = "contrast",
                                            toscale = c("pval"),
                                            fcestimate = "diff",
                                            benchmark = list(
                                              list(score = "diff", desc = TRUE),
                                              list(score = "t_statistic", desc = TRUE),
                                              list(score = "scaled.pval", desc = TRUE)
                                            ),  
                                            model_description = "proDA_medpolishInt",
                                            model_name = "proDA_medpolishInt",
                                            FDRvsFDP = list(list(score = "adj_pval", desc = FALSE))
                                            , hierarchy = c("name"), summarizeNA = "t_statistic"
)

sumarry <- benchmark_proDA$smc$summary
prolfqua::table_facade(sumarry, caption = "nr of proteins with 0, 1, 2, 3 missing contrasts.")

```

```{r benchmarkProDA}
res <- benchmark_proDA$pAUC_summaries()
knitr::kable(res$ftable$content,caption = res$ftable$caption)
res$barp
```



# Fit model with msqrob

```{r setUPQFeatures}
se <- prolfqua::LFQDataToSummarizedExperiment(lfqtransformed)

pe <- QFeatures::QFeatures(list(peptide = se), colData = colData(se))

my_medianPolish <- function(x, verbose = FALSE, ...){
  medpol <- stats::medpolish(x, na.rm = TRUE, trace.iter = verbose, maxiter = 10)
  return(medpol$overall + medpol$col)
}


pe <- QFeatures::aggregateFeatures(
  pe,
  i = "peptide", fcol = "protein_Id",
  name = "protein", fun = my_medianPolish, 
)

```



```{r hurdleModel}
library(msqrob2)
prlm <- msqrobHurdle(pe,
                     i = "protein",
                     formula = ~group.,
                     overwrite = TRUE)

L <- makeContrast(c("group.b=0"),
                  parameterNames = c("group.b"))
prlm <- hypothesisTestHurdle(prlm, i = "protein", L, overwrite = TRUE)

xx <- rowData(prlm[["protein"]])
hurdle <- xx[grepl("hurdle_",names(xx))]

res <- list()
for (i in names(hurdle)) { 
  hurdle[[i]]$contrast <- i
  res[[i]] <- prolfqua::matrix_to_tibble(hurdle[[i]], preserve_row_names = "name")
}

hurdle <- dplyr::bind_rows(res)

logFC <- hurdle |> dplyr::select("name","contrast", starts_with("logFC"))
logFC <- filter(logFC ,!is.na(logFCt))
logFC$modelName <- "msqrobHurdleIntensity"
names(logFC) <- c("name","contrast","logFC","se","df","t","pval","modelName")

logOR <- hurdle |> dplyr::select("name","contrast", starts_with("logOR"))
logOR$modelName <- "msqrobHurdleCount"
names(logOR) <- c("name","contrast","logFC","se","df","t","pval","modelName")
ddd <- dplyr::anti_join(logOR , logFC, by = c("name", "contrast"))
all <- dplyr::bind_rows(ddd , logFC) |> dplyr::arrange(contrast, name)
all <- prolfqua::adjust_p_values(all, column = "pval", group_by_col = "contrast")
all$contrast <- "b_vs_a"


st <- lfqProt$get_Stats()
protAbundanceIngroup <- st$stats()

protAbundanceIngroup <- protAbundanceIngroup |> 
  tidyr::pivot_wider(id_cols = protein_Id,
                     names_from = group., names_prefix = "abd.", 
                     values_from = mean)
protAbundanceIngroup <- protAbundanceIngroup |> dplyr::mutate( avgAbd.b.a =  mean( c(abd.b,abd.a), na.rm = TRUE)  )
protAbundanceIngroup <- protAbundanceIngroup |> 
  dplyr::select(-starts_with("abd")) |> 
  tidyr::pivot_longer(starts_with("avgAbd"), names_to = "contrast" ,values_to = "avgAbd")

protAbundanceIngroup$contrast <- "b_vs_a" 
bb <- dplyr::inner_join(all, protAbundanceIngroup, by = c("name" = "protein_Id", "contrast" = "contrast"))

```

```{r}


ttd <- prolfquabenchmark::cptac_bench_preprocess(bb, idcol = "name")
benchmark_msqrob <- prolfqua::make_benchmark(ttd$data,
                                             contrast = "contrast",
                                             toscale = c("pval"),
                                             fcestimate = "logFC",
                                             benchmark = list(
                                               list(score = "logFC", desc = TRUE),
                                               list(score = "t", desc = TRUE),
                                               list(score = "scaled.pval", desc = TRUE)
                                             ),  
                                             model_description = "msqrob2_QFeature",
                                             model_name = "msqrob2_QFeature",
                                             FDRvsFDP = list(list(score = "pval.adjusted", desc = FALSE))
                                             , hierarchy = c("name"), summarizeNA = "t"
)

sum(benchmark_msqrob$smc$summary$name)
sumarry <- benchmark_msqrob$smc$summary
prolfqua::table_facade(sumarry, caption = "nr of proteins with 0, 1, 2, 3 missing contrasts.")

```

```{r}
res <- benchmark_msqrob$pAUC_summaries()
knitr::kable(res$ftable$content,caption = res$ftable$caption)
res$barp
```


# Compare benchmark results.

```{r fig.cap="Comparing msqrob2, proda and prolfqua using the pAUC_{10}"}
saveRDS(list(benchmark_msqrob = benchmark_msqrob,
             benchmark_proDA = benchmark_proDA,
             benchmark_merged = benchmark_merged),
        file = "../inst/Benchresults/CPTAC_Benchmark.RDS")

mrob <- benchmark_msqrob$pAUC_summaries()$ftable$content
mrob$package <- "msqrob2"
proda <- benchmark_proDA$pAUC_summaries()$ftable$content
proda$package <- "proDA"
prolfqua <- benchmark_merged$pAUC_summaries()$ftable$content
prolfqua$package <- "prolfqua"

all <- dplyr::bind_rows(list(mrob, proda, prolfqua))
all <- all |> filter(contrast == "b_vs_a")
all$what[all$what == "statistic"] <- "t_statistic"
all$what[all$what == "t"] <- "t_statistic"
all$what[all$what == "scaled.pval"] <- "scaled.p.value"
all$what[all$what == "logFC"] <- "diff"

all |> ggplot(aes(x = what, y = pAUC_10, group = package, color = NULL, fill = package)) +
  geom_bar(stat = "identity",  position = position_dodge()) 

```



```{r roctest, include = TRUE, message=FALSE, error=FALSE}

ttmsqrob <- benchmark_msqrob$data()
ddmsqrob <- pROC::roc(ttmsqrob$TP, ttmsqrob$scaled.pval, partial.auc = c(1, 0.9))

ttprolfqua <- benchmark_merged$data()
ddprolfqua <- pROC::roc(ttprolfqua$TP, ttprolfqua$scaled.p.value, partial.auc = c(1, 0.9))

ttproda <- benchmark_proDA$data()
ddproda <- pROC::roc(ttproda$TP, ttproda$scaled.pval, partial.auc = c(1, 0.9))


tmp <- c(msqrob2_vs_prolfqua = pROC::roc.test(ddmsqrob,ddprolfqua, progress = "none")$p.value,
         msqrob2_vs_proda = pROC::roc.test(ddmsqrob,ddproda, progress = "none")$p.value,
         prolfqua_vs_proda = pROC::roc.test(ddprolfqua,ddproda, progress = "none")$p.value)

```

We observe that for this benchmark data, there are no significant differences among the $AUC_{10}$ for the three packages. The Table \@ref(tab:shwoROCtestREsults) shows the results of the Bootstrap test for two ROC curves.

```{r shwoROCtestREsults}
knitr::kable(data.frame(pROC_test = names(tmp), p.value = tmp), caption = "p-values for pairwise comparsions of partial AUC")

```
