---
title: "Benchmarking proDA and prolfqua using Ionstar Dataset and MSFragger IonQuant compbined_protein output"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Benchmarking proDA and prolfqua using Ionstar Dataset and MSFragger IonQuant compbined_protein output} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

Please download and install the `prolfquadata` package from github

```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo = TRUE,
message = FALSE,
warning = FALSE,
fig.width = 10,
fig.height = 10
)
```

```{r conflicted}
conflicted::conflict_prefer("filter", "dplyr")

```

## Load data

We start by loading the IonStar dataset and the annotation from the `prolfquadata` package. The method `add_annotation` adds the annotation to the data.



```{r loaddata}
datadir <- file.path(find.package("prolfquadata") , "quantdata")
inputFragfile <-  file.path(datadir, "MSFragger_IonStar2018_PXD003881.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")
annotation <- readxl::read_xlsx(inputAnnotation)

protein <- tibble::as_tibble(read.csv(unz(inputFragfile,"IonstarWithMSFragger/combined_protein.tsv"),
                             header = TRUE, sep = "\t", stringsAsFactors = FALSE))

undebug( prolfqua::tidy_FragPipe_combined_protein)
protein <- prolfqua::tidy_FragPipe_combined_protein_deprec(protein)

protein <- protein |> dplyr::filter(unique.stripped.peptides > 1)
merged <- dplyr::inner_join(annotation, protein)
```

## Create prolfqua configuration

```{r setupConfigs}

atable <- prolfqua::AnalysisTableAnnotation$new()
atable$fileName = "raw.file"
atable$hierarchy[["protein_Id"]] <- c("protein")

atable$hierarchyDepth <- 1
atable$set_response("total.intensity")
atable$factors[["dilution."]] = "sample"
atable$factors[["run"]] = "run_ID"
atable$factorDepth <- 1


config <- prolfqua::AnalysisConfiguration$new(atable)

adata <- prolfqua::setup_analysis(merged, config)
lfqdata <- prolfqua::LFQData$new(adata, config)
lfqdata$remove_small_intensities()


```

## Normalize data using human proteins

```{r createPlotter}
pl <- lfqdata$get_Plotter()
pl$intensity_distribution_density()
subset_h <- lfqdata$get_copy()$get_Transformer()$log2()$lfq
subset_h$data <- subset_h$data |> dplyr::filter(grepl("HUMAN", protein_Id))
tr <- lfqdata$get_Transformer()
lfqdataNormalized <- tr$log2()$robscale_subset(lfqsubset = subset_h)$lfq
pl <- lfqdataNormalized$get_Plotter()
pl$intensity_distribution_density()
hm <- pl$NA_heatmap()
```

```{r showHeatmap}
hm
```



# Model data using prolfqua and specify contrasts

```{r specifyContrasts}
Contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c",
  "dilution_(6/4.5)_1.3(3)" =   "dilution.c - dilution.b",
  "dilution_(4.5/3)_1.5" =   "dilution.b - dilution.a"
)


lmmodel <- "~ dilution."
lmmodel <- paste0(lfqdataNormalized$config$table$get_response() , lmmodel)

modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "Model")

mod <- prolfqua::build_model(lfqdataNormalized$data, modelFunction)
contr <- prolfqua::Contrasts$new(mod, Contrasts)
contrimp <- prolfqua::ContrastsSimpleImpute$new(lfqdataNormalized, Contrasts)

merged <- prolfqua::merge_contrasts_results(contr, contrimp)
mergedmod <- prolfqua::ContrastsModerated$new(merged$merged)

cp <- mergedmod$get_Plotter()
cp$volcano()

```


## Benchmark data 


```{r benchmark}
ttd <- prolfqua::ionstar_bench_preprocess(mergedmod$get_contrasts())
medpol_benchmark <- prolfqua::make_benchmark(ttd$data,
                                   model_description = "MSFragger med. polish and lm. density",
                                   model_name = "MSFragger_prot_med_lm",
                                   FDRvsFDP = list(list(score = "FDR", desc = FALSE))
)

#saveRDS(medpol_benchmark, "../inst/Benchresults/MSFragger_medpol_benchmark.RDS")

knitr::kable(medpol_benchmark$pAUC_summaries()$ftable$content)


prolfqua::table_facade(medpol_benchmark$smc$summary, "Nr of estimated contrasts")
medpol_benchmark$plot_score_distribution()
medpol_benchmark$plot_ROC(0.05)
medpol_benchmark$plot_FDRvsFDP()

```


# Model data using proDA and specify contrasts

```{r prodaFit}
# Not Run
se <- prolfqua::LFQDataToSummarizedExperiment(lfqdataNormalized)
fit <- proDA(se, design = ~ dilution. - 1, data_is_log_transformed = TRUE)

contr <- list()
contr[["dilution_(9/7.5)_1.2"]] <- data.frame(contrast = "dilution_(9/7.5)_1.2",
                                              test_diff(fit, contrast = "dilution.e - dilution.d"))
contr[["dilution_(7.5/6)_1.25"]] <- data.frame(contrast = "dilution_(7.5/6)_1.25",
                                               test_diff(fit, contrast = "dilution.d - dilution.c"))
contr[["dilution_(6/4.5)_1.3(3)"]] <- data.frame(contrast = "dilution_(6/4.5)_1.3(3)", 
                                                 test_diff(fit, contrast = "dilution.c - dilution.b"))
contr[["dilution_(4.5/3)_1.5"]] <- data.frame(contrast = "dilution_(4.5/3)_1.5", 
                                              test_diff(fit, contrast = "dilution.b - dilution.a" ))

bb <- dplyr::bind_rows(contr)
# End Not Run


```


## Benchmark data 


```{r prodabenchmark}

ttd <- prolfqua::ionstar_bench_preprocess( bb , idcol = "name" )

benchmark_proDA <- prolfqua::make_benchmark(ttd$data,
                                   contrast = "contrast",
                                toscale = c("pval"),
                                fcestimate = "diff",
                                benchmark = list(
                                  list(score = "diff", desc = TRUE),
                                  list(score = "t_statistic", desc = TRUE),
                                  list(score = "scaled.pval", desc = TRUE)
                                ),  
                                model_description = "proDA_medpolishInt",
                                model_name = "proDA_medpolishInt",
                                FDRvsFDP = list(list(score = "adj_pval", desc = FALSE))
, hierarchy = c("name"), summarizeNA = "t_statistic"
)

sumarry <- benchmark_proDA$smc$summary
prolfqua::table_facade(sumarry, caption = "nr of proteins with 0, 1, 2, 3 missing contrasts.")

knitr::kable(benchmark_proDA$pAUC_summaries()$ftable$content)

prolfqua::table_facade(medpol_benchmark$smc$summary, "Nr of estimated contrasts")

benchmark_proDA$plot_score_distribution()
benchmark_proDA$plot_ROC(0.05)
benchmark_proDA$plot_FDRvsFDP()

```


# Compare prolfqua and proda

Direct comparison with msqrob2 is not possible since to fit the dropout model also the peptide intensities are required.

```{r}

proda <- benchmark_proDA$pAUC_summaries()$ftable$content
proda$package <- "proda"
prolfqua <- medpol_benchmark$pAUC_summaries()$ftable$content
prolfqua$package <- "prolfqua"

tmp <- bind_rows(proda, prolfqua)
tmp$what |> unique()
tmp$what[tmp$what == "statistic"] <- "t_statistic"
tmp$what[tmp$what == "scaled.pval"] <- "scaled.p.value"

tmp |> ggplot(aes(x = what, y= pAUC_10, group = package, color = NULL, fill = package)) + geom_bar(stat = "identity",  position = position_dodge()) + facet_wrap(~ contrast) +  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```




