---
title: "Benchmarking normalization, aggregation and models using the Ionstar Dataset"
author: "FGCZ - (Draft)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      toc_collapsed: false
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Benchmarking normalization, aggregation and models using the Ionstar Dataset} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

Please download and install the `prolfquadata` package from github

```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo = TRUE,
message = FALSE,
warning = FALSE,
fig.width = 5,
fig.height = 5,
knitr.table.format = "pipe"
)
```

```{r loadRequiredPackages}
conflicted::conflict_prefer("filter", "dplyr")

```

Decide if you work with all data or for speedup with subset of data:

```{r}
SUBSET <- FALSE
SUBSETNORM <- TRUE
SAVE <- TRUE

```

We start by loading the IonStar dataset and the annotation from the `prolfquadata` package.
The method `add_annotation` adds the annotation to the data.

```{r loadMQData}

datadir <- file.path(find.package("prolfquadata") , "quantdata")
inputMQfile <-  file.path(datadir,
                          "MAXQuant_IonStar2018_PXD003881.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")
mqdata <- list()

mqdata$data <- prolfqua::tidyMQ_Peptides(inputMQfile)
length(unique(mqdata$data$proteins))

mqdata$config <- prolfqua::create_config_MQ_peptide()


annotation <- readxl::read_xlsx(inputAnnotation)
res <- prolfqua::add_annotation(
  mqdata$data,
  annotation,
  fileName = "raw.file"
)

```

The `setup_analysis` asserts that all columns specified in the configruation are present in the data.
For more details about the `prolfqua` configuration see the vignette "Creating Configurations".

```{r annotateData}
mqdata$config$table$factors[["dilution."]] = "sample"
mqdata$config$table$factors[["run_Id"]] = "run_ID"
mqdata$config$table$factorDepth <- 1
mqdata$data <- prolfqua::setup_analysis(res, mqdata$config)

```

## Data normalization

First we remove all contaminant, decoy proteins from the list, than we remove 0 intensity values, then filter for 2 peptides per protein.

```{r filterForHumans}
lfqdata <- prolfqua::LFQData$new(mqdata$data, mqdata$config)
lfqdata$data <- lfqdata$data |> dplyr::filter(!grepl("^REV__|^CON__", protein_Id)) 
sr <- lfqdata$get_Summariser()
lfqdata$remove_small_intensities()
sr <- lfqdata$get_Summariser()
sr$hierarchy_counts()

```

We will normalize the data using the 'LFQTransformer' class.
Since we know that the Human proteins are the Matrix in the experiment we will normalize the data using HUMAN proteins only.
To this task we subset the dataset by filtering for HUMAN proteins only and then use the `LFQDataTransformer` to normalize the data.

```{r}
tr <- lfqdata$get_Transformer()
subset_h <- lfqdata$get_copy()
subset_h$data <- subset_h$data |> dplyr::filter(grepl("HUMAN", protein_Id))
subset_h <- subset_h$get_Transformer()$log2()$lfq
lfqdataNormalized <- tr$log2()$robscale_subset(lfqsubset = subset_h,  preserveMean = FALSE )$lfq

```

The figures below show the intensity distribution before and after normalization.

```{r plotIntensityDistributions}

before <- lfqdata$get_Plotter()
before$intensity_distribution_density()

after <- lfqdataNormalized$get_Plotter()
after$intensity_distribution_density()

```

Create a sample of N proteins to speed up computations of models and contrasts.

```{r createDataSubset}
if (SUBSET) {
  N <- 200
  mqdataSubset <- lfqdata$get_sample(size = N, seed = 2020)
  lfqNormSubset <- lfqdataNormalized$get_sample(size = N, seed = 2020)
  lfqNormSubset$hierarchy_counts()
} else {
  mqdataSubset <- lfqdata$get_copy()  
  lfqNormSubset <- lfqdataNormalized$clone()
  lfqNormSubset$hierarchy_counts()
}
```



```{r R6Ionstar, echo =  FALSE, include = FALSE}

IonstarData <- R6::R6Class(
  "IonstarData",
  public = list(
    data = NULL,
    config = NULL,
    data_N = NULL,
    config_N = NULL,
    
    initialize = function(data, config, data_N, config_N){
      self$data = data
      self$config = config
      self$data_N = data_N
      self$config_N = config_N
    },
    Pep = function(){
      return(list(data = self$data, config = self$config$clone(deep = TRUE)))
    },
    filtered = function(){
      data <- prolfqua:::filter_proteins_by_peptide_count( self$data ,  self$config )$data
      return(list(data = data, config = self$config$clone(deep = TRUE)))
    },
    normalized = function(){
      return(list(data = self$data_N, config = self$config_N))
    },
    subset_normalized = function(){
      return(list(data = self$data_N, config = self$config_N))
    }
  )
)

ionstar <- IonstarData$new(mqdataSubset$data, mqdataSubset$config, 
                           lfqNormSubset$data, lfqNormSubset$config)

prolfqua::nr_B_in_A( mqdataSubset$data ,  mqdataSubset$config )

#usethis::use_data(ionstar, overwrite = TRUE)
ionstar$Pep()
ionstar$filtered()
ionstar$normalized()
ionstar$subset_normalized()

```

# Inferring Protein abundances from Peptide abundances

We will be using the `LFQDataAggregator` class.
To estimate protein abundances using Tukey's median polish we need to use log2 transformed peptide abundances

```{r aggregateMedpolish, fig.width=10, fig.height=10}
lfqNormSubset$filter_proteins_by_peptide_count()

lfqNormSubset$config$table$get_response()

pl <- lfqNormSubset$get_Plotter()
pl$intensity_distribution_density()
lfqAggMedpol <- lfqNormSubset$get_Aggregator()
lfqAggMedpol$medpolish()
```

The figure below shows the the peptide abundances used for estimation and the protein abundance estimates (black line).

```{r}
xx <- lfqAggMedpol$plot()
gridExtra::grid.arrange(grobs =  xx$plots[1:3])
```

```{r storeProteinAggregates, echo = FALSE, include = FALSE}
data_IonstarProtein_subsetNorm <- list(
  data = lfqAggMedpol$lfq_agg$data,
  config = lfqAggMedpol$lfq_agg$config
)
```

We can also estimate the protein abundances by adding the abundances of the top N most abundant peptides.
In this case we are using the untransformed peptide abudances.

```{r topNAggregation, echo = TRUE, eval=FALSE}
lfqAggregator <- lfqdata$get_Aggregator()
lfqAggregator$mean_topN()
topN <- lfqAggregator$plot()
topN$plots[[1]]

```

# Model Fitting

We will be fitting tree models to the data:

- The first model is a linear model as implemented by the R function `lm` fitted to protein abundances inferred from peptide abundances using the LFQAggregator.
- The second model is mixed effects model implemented in the R function `lmer` fitted to peptide abundances, where we model the peptide measurements as repeated measurements of the protein.
- The third model again is a linear model but fitted to peptide abundances. By this we obtain for each peptide a linear model can compute contrasts and then can aggregate them using the ROPECA method.

## Fitting a linear model to the protein abundances

```{r buildModelLM}
protLFQ <- lfqAggMedpol$lfq_agg
sr <- protLFQ$get_Summariser()
sr$hierarchy_counts()

lmmodel <- "~ dilution."
lmmodel <- paste0(protLFQ$config$table$get_response() , lmmodel)

lfqNormSubset$config$table$hierarchyDepth <- 1
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "Model")

modLinearProt <- prolfqua::build_model(protLFQ$data, modelFunction)
modLinearProt$anova_histogram()$plot

```



## Fitting a mixed effects model to peptide abundances

```{r buildModelLMer}

lmmodel <- "~ dilution. + (1|peptide_Id) + (1|sampleName)"
lmmodel <- paste0(lfqNormSubset$config$table$get_response() , lmmodel)
lfqNormSubset$config$table$hierarchyDepth <- 1
modelFunction <- prolfqua::strategy_lmer( lmmodel, model_name = "Model")
modMixedProtLevel <- prolfqua::build_model(lfqNormSubset$data, modelFunction)
modMixedProtLevel$anova_histogram()$plot

```



## Fitting peptide level models

```{r buildModelPeptide}

lmmodel <- "~ dilution."

lfqNormSubset$config$table$hierarchyDepth
lfqNormSubset$config$table$hierarchyDepth <- 2

lmmodel <- paste0(lfqNormSubset$config$table$get_response() , lmmodel)

modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "Model")
modLMPepLevel <- prolfqua::build_model(lfqNormSubset$data,
                                       modelFunction,
                                       subject_Id = lfqNormSubset$subject_Id())
modLMPepLevel$anova_histogram()$plot

```

# Computing Contrasts

Once models are fitted contrasts can be computed.
The R code below defines all possible contrasts among conditions for the ionstar dataset.

```{r defineContrasts}
DEBUG <- FALSE

Contrasts <- c(
  "dilution_(9/3)_3" =   "dilution.e - dilution.a",
  "dilution_(9/4.5)_2" =   "dilution.e - dilution.b",
  "dilution_(9/6)_1.5" =   "dilution.e - dilution.c",
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  
  "dilution_(7.5/3)_2.5" =   "dilution.d - dilution.a",
  "dilution_(7.5/4.5)_1.6(6)" =   "dilution.d - dilution.b",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c",
  
  "dilution_(6/3)_2" =   "dilution.c - dilution.a",
  "dilution_(6/4.5)_1.3(3)" =   "dilution.c - dilution.b",
  
  "dilution_(4.5/3)_1.5" =   "dilution.b - dilution.a"
)


tt <- Reduce(rbind, strsplit(names(Contrasts),split = "_"))
tt <- data.frame(tt)[,2:3]
colnames(tt) <- c("ratio" , "expected fold-change")
tt <- tibble::add_column(tt, contrast =  Contrasts, .before = 1)
prolfqua::table_facade(
  tt,
  caption = "All possible Contrasts given 5 E. coli dilutions of the Ionstar Dataset", digits = 1)

```

```{r usedContrasts}
relevantContrasts <- c("dilution_(9/7.5)_1.2",
                       "dilution_(7.5/6)_1.25",
                       "dilution_(6/4.5)_1.3(3)",
                       "dilution_(4.5/3)_1.5" )

tt <- Reduce(rbind, strsplit(relevantContrasts,split = "_"))
tt <- data.frame(tt)[,2:3]
colnames(tt) <- c("ratio" , "expected fold-change")
tt <- tibble::add_column(tt, contrast =  Contrasts[names(Contrasts) %in% relevantContrasts], .before = 1)
prolfqua::table_facade(tt, caption = "Contrasts used for benchmark.", digits = 1)
relevantContrasts <- Contrasts[names(Contrasts) %in% relevantContrasts]

```

There are, as of today, four contrasts classes in the package *prolfqua*:

- 'ContrastsSimpleImputed' : contrast computation with imputation of fold changes and t-statistic estimation using pooled variances.
- 'Contrasts' : uses Wald test, 
- 'ContrastsModerated' : applies variance moderation,
- 'ContrastsROPECA' implements difference and p-value aggregation

## Contrasts with Imputation

In order to estimate differences (fold-changes), statistics and p-values of proteins for which linear models could not be fitted because of an excess of missing measurements, the following procedure is applied.
The mean abundance of a protein in a condition is computed.
For the proteins with no observation in a condition, we infer their abundances by using the mean protein abundances observed only in one sample per group.
The standard deviation of the protein is estimated using the pooled variances of the condition where the variance could be estimated.

```{r contrastImputed}
contrImp <- prolfqua::ContrastsSimpleImpute$new(
  protLFQ,
  relevantContrasts)

bb <- contrImp$get_contrasts()

plc <- contrImp$get_Plotter()
plc$volcano()$FDR
plc$ma_plot()
plc$histogram()$p.value

```



```{r benchmarkImputed}
allContrasts <- list()
allContrasts$imputation <- contrImp$get_contrasts()
ttd <- prolfqua::ionstar_bench_preprocess(contrImp$get_contrasts())

benchmark_imputation <- prolfqua::make_benchmark(
    ttd$data,
    model_description = "med. polish and lm. imputation",
    model_name = "prolfqua_impute",
    FDRvsFDP = list(list(score = "FDR", desc = FALSE))
)

benchmark_imputation$plot_ROC(xlim = 0.1)


benchmark_imputation$plot_FDRvsFDP()

prolfqua::table_facade(benchmark_imputation$smc$summary, caption = "Nr of proteins with Nr of not estimated contrasts.", digits = 1)
```

```{r fig.cap = "Summary of partial area under the ROC curve."}

bb <- benchmark_imputation$pAUC_summaries()
bb$barp
```


```{r}
prolfqua::table_facade(bb$ftable$content, caption = bb$ftable$caption, digits = 1)

allBenchmarks <- list()
allBenchmarks$benchmark_imputation <- benchmark_imputation

```


## Contrasts from linear model

```{r contrastModel1}

contrProt <- prolfqua::Contrasts$new(modLinearProt, relevantContrasts)
pl <- contrProt$get_Plotter()
pl$volcano()$FDR
pl$histogram()$p.value

```

```{r benchmarkModel1}
allContrasts$Prot <- contrProt$get_contrasts()

ttd <- prolfqua::ionstar_bench_preprocess(contrProt$get_contrasts())

benchmark_Prot <- prolfqua::make_benchmark(
    ttd$data,
    model_description = "med. polish and lm",
    model_name = "prolfqua_lm"
)

prolfqua::table_facade(benchmark_Prot$smc$summary, caption = "Nr of proteins with Nr of not estimated contrasts.", digits = 1)

#benchmark_Prot$plot_score_distribution()
benchmark_Prot$plot_FDRvsFDP()
allBenchmarks$benchmark_Prot <- benchmark_Prot

```


### Adding Moderation

```{r contrastModel1Moderation, include = FALSE}

contrProtModerated <- prolfqua::ContrastsModerated$new(contrProt)
contrProtModerated$get_Plotter()$volcano()$FDR

```

```{r benchmarkModel1Moderation, include = FALSE}
allContrasts$ProtModerated <- contrProtModerated$get_contrasts()
ttd <- prolfqua::ionstar_bench_preprocess(contrProtModerated$get_contrasts())
benchmark_ProtModerated <- prolfqua::make_benchmark(
    ttd$data,
    model_description = "med. polish and lm moderated",
    model_name = "prolfqua_lm_mod")

prolfqua::table_facade(
  benchmark_ProtModerated$smc$summary,
  caption = "Nr of proteins with Nr of not estimated contrasts.",
  digits = 1)

benchmark_ProtModerated$plot_FDRvsFDP()
allBenchmarks$benchmark_ProtModerated <- benchmark_ProtModerated

```



## Contrasts from mixed effect models

```{r contrastModel2}
contrProtMixed <- prolfqua::Contrasts$new(modMixedProtLevel, relevantContrasts, modelName = "WaldTestMixed")

pl <- contrProtMixed$get_Plotter()
pl$volcano()$FDR
pl$histogram()$p.value

pl <- contrProtMixed$get_contrasts()
pl$protein_Id |> unique() |> length()
```

```{r benchmarkModel2}
allContrasts$contrProtMixed <- contrProtMixed$get_contrasts()
ttd <- prolfqua::ionstar_bench_preprocess(contrProtMixed$get_contrasts())
benchmark_mixed <- prolfqua::make_benchmark(
    ttd$data,
    model_description = "mixed effect model",
    model_name = "prolfqua_mix_eff"
)
benchmark_mixed$complete(FALSE)
prolfqua::table_facade(benchmark_mixed$smc$summary,
                       caption = "Nr of proteins with Nr of not estimated contrasts.", digits=1)

#benchmark_mixed$plot_score_distribution()
benchmark_mixed$plot_FDRvsFDP()
allBenchmarks$benchmark_mixed <- benchmark_mixed
```

### Adding Moderation

Since moderation requires a degrees of freedom estimate to determine the prior degrees of freedom we examine the denominator degrees of freedom produced by the methods implemented in lmerTest (see Histogram).

```{r distributionOfDFModel2, fig.cap="Histogram of degrees of freedom for mixed model"}
ctr <- contrProtMixed$get_contrasts()
df <- ctr$df
df[df > 59] <- 60
range(df)
hist(df, breaks = 100, xlim = c(0,61))
```

```{r contrastModel2Moderation}
contrProtMixedModerated <- prolfqua::ContrastsModerated$new(contrProtMixed)
contrProtMixedModerated$get_Plotter()$volcano()$FDR

```

```{r benchmarkModel2Moderation}
allContrasts$contrProtMixedModerated <- contrProtMixedModerated$get_contrasts()

ttd <- prolfqua::ionstar_bench_preprocess(contrProtMixedModerated$get_contrasts())

benchmark_mixedModerated <- prolfqua::make_benchmark(
    ttd$data,
    model_description = "mixed effect model moderated",
    model_name = "prolfqua_mix_eff_mod")
prolfqua::table_facade(benchmark_mixedModerated$smc$summary, caption = "Nr of proteins with Nr of computed contrasts.", digits=1)

#benchmark_mixedModerated$plot_score_distribution()
benchmark_mixedModerated$plot_FDRvsFDP()
allBenchmarks$benchmark_mixedModerated <- benchmark_mixedModerated
```

## Protein level contrasts from peptide models

To estimate regulation probabilities using the ROPECA approach we can chain the contrast computation methods.
First we compute contrasts on peptide level, than we moderated the variance, t-statistics and p-values and finally we aggregate the fold change estimates and p-values.

```{r contrastROPECA}
contrROPECA <- prolfqua::Contrasts$new(modLMPepLevel, relevantContrasts) |>
  prolfqua::ContrastsModerated$new() |>
  prolfqua::ContrastsROPECA$new()

contrROPECA$get_Plotter()$volcano()$FDR
contrROPECA$get_Plotter()$histogram()$FDR

```

```{r benchmarkROPECA}
cr <- contrROPECA$get_contrasts()
ttd <- prolfqua::ionstar_bench_preprocess(cr)
benchmark_ropeca <- prolfqua::make_benchmark(
    ttd$data,
    toscale = c("beta.based.significance"),
    benchmark = list(
        list(score = "diff", desc = TRUE),
        list(score = "statistic", desc = TRUE),
        list(score = "scaled.beta.based.significance", desc = TRUE)
    ),  
    model_description = "Ropeca",
    model_name = "prolfqua_ropeca",
    FDRvsFDP = list(list(score = "FDR.beta.based.significance", desc = FALSE))
)

prolfqua::table_facade(
    benchmark_ropeca$smc$summary,
    caption = "Nr of proteins with Nr of not estimated contrasts.",
    digits = 1)

benchmark_ropeca$plot_ROC(0.1)
benchmark_ropeca$plot_FDRvsFDP()

allBenchmarks$benchmark_ropeca <- benchmark_ropeca

```

```{r fig.cap = "Summary of partial area under the ROC curve."}
bb <- benchmark_ropeca$pAUC_summaries()
bb$barp
```



```{r}
prolfqua::table_facade(bb$ftable$content, caption = bb$ftable$caption, digits=1)

```

## Merging contrasts

Here we merge contrasts estimates from linear models and from the models with imputation.

```{r mergingContrasts}
all <- prolfqua::merge_contrasts_results(prefer = contrProt, add = contrImp)

merged <- ContrastsModerated$new(all$merged)
ttd <- prolfqua::ionstar_bench_preprocess(merged$get_contrasts())
benchmark_merged <- prolfqua::make_benchmark(
    ttd$data,
    model_description = "merge of prot moderated and imputed",
    model_name = "prolfqua_merged")

prolfqua::table_facade(benchmark_merged$smc$summary, caption = "Nr of proteins with Nr of not estimated contrasts.", digits=1)
#benchmark_mixedModerated$plot_score_distribution()

benchmark_merged$plot_FDRvsFDP()

```

```{r fig.width=10, fig.height=10, fig.cap="ROC curves for merged benchmark"}
benchmark_merged$plot_ROC(xlim = 0.15)

bb <- benchmark_merged$pAUC_summaries()
bb$barp

tmp <- prolfqua::table_facade(bb$ftable$content, caption = bb$ftable$caption, digits=1)
knitr::kable((bb$ftable$content))

allBenchmarks$benchmark_merged <- benchmark_merged


```

```{r same, eval=FALSE}
same <- all$same
allBenchmarks$benchmark_Prot$smc$summary

ttd <- prolfqua::ionstar_bench_preprocess(same$get_contrasts())
benchmark_same <- prolfqua::make_benchmark(
    ttd$data,
    model_description = "imputed_same_as_lm",
    model_name = "imputed_same_as_lm")

prolfqua::table_facade(benchmark_same$smc$summary, caption = "Nr of proteins with Nr of not estimated contrasts.", digits=1)
benchmark_same$plot_FDRvsFDP()

```

```{r storeSame, fig.width=10, fig.height=10, fig.cap="ROC curves for merged benchmark", eval=FALSE}
benchmark_same$plot_ROC(xlim = 0.15)
bb <- benchmark_same$pAUC_summaries()
bb$barp

prolfqua::table_facade(bb$ftable$content, caption = bb$ftable$caption, digits=1)

allBenchmarks$benchmark_same <- benchmark_same


```

```{r include=FALSE}

if(SAVE){
    if (SUBSETNORM) {
        saveRDS(allBenchmarks, file = "../inst/Benchresults/allBenchmarks.RDS")
        saveRDS(allContrasts, file = "../inst/Benchresults/allContrasts.RDS")
    } else{
        saveRDS(allContrasts, file = "../inst/Benchresults/allContrastsNORMAL.RDS")
        saveRDS(allBenchmarks, file = "../inst/Benchresults/allBenchmarksNORMAL.RDS")
    }
    
    
}

```

# Comparing various models

The table below summarizes the contrast estimates produced which will be benchmarked.

|                          |      Model      | Contrast | Moderation | Aggregation |
|--------------------------|:---------------:|----------|------------|-------------|
| Protein abudance         |       lm        | o        | o          |             |
| Protein abudance Imputed | pooled variance | o        | o          |             |
| Peptide abudance         |      lmer       | o        | o          |             |
| Peptide abudance         |       lm        |          |            | o           |

```{r fig.cap="Partial area under the ROC curve at 10% FPR."}
ttt <- sapply(allBenchmarks, function(x){x$complete(FALSE)})
res <- purrr::map_df(allBenchmarks, function(x){x$pAUC()})
resAllB <- res |> dplyr::filter(contrast == "all")

bb <- resAllB |> dplyr::mutate(whatfix = dplyr::case_when(what == "scaled.beta.based.significance" ~ "scaled.p.value", TRUE ~ what))

ggplot2::ggplot(bb, ggplot2::aes(x = Name, y = pAUC_10)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::facet_wrap(~whatfix)  + 
  ggplot2::coord_cartesian(ylim = c(min(bb$pAUC_10),max(bb$pAUC_10))) + 
  ggplot2::theme_minimal() + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5))

```

```{r fig.caption="Partial area under the ROC curve at 20% FPR."}
ggplot2::ggplot(bb, ggplot2::aes(x = Name, y = pAUC_20)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::facet_wrap(~whatfix)  +
  ggplot2::coord_cartesian(ylim = c(min(bb$pAUC_20),max(bb$pAUC_20))) + 
  ggplot2::theme_minimal() + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5))  

```

### Look at the nr of estimated contrasts.

```{r fig.cap="NR of estimated contrasts", fig.width=5, fig.height=5}
dd <- purrr::map_df(allBenchmarks, function(x){res <- x$smc$summary; res$name <- x$model_name;res})


dd <- dd |> dplyr::mutate(nrcontrasts = protein_Id * (4 - as.integer(nr_missing)))
dds <- dd |> dplyr::group_by(name) |> dplyr::summarize(nrcontrasts = sum(nrcontrasts))

dds |> ggplot2::ggplot(ggplot2::aes(x = name, y = (nrcontrasts - min(nrcontrasts)))) + 
  ggplot2::geom_bar(stat="identity") + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::geom_text(ggplot2::aes(label= nrcontrasts), position = ggplot2::position_dodge(width=0.9), vjust=-0.25)

```

### Plot FDR vs FDP

```{r FDRfdp, fig.cap = "Compare FDR estimate with false discovery proportion (FDP).", fig.width=8, fig.height=8}
dd <- purrr::map_df(allBenchmarks, function(x){res <- x$get_confusion_FDRvsFDP(); res$name <- x$model_name;res})

dd |> ggplot2::ggplot(ggplot2::aes(y = FDP_,  x  = scorecol )) + 
  ggplot2::geom_line(ggplot2::aes(color = model_name)) +
  ggplot2::facet_wrap(~contrast) + 
  ggplot2::geom_abline(intercept = 0, slope = 1, color = 2)

```
