---
title: "Modelling with Interactions"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
vignette: |
  %\VignetteIndexEntry{Modelling with Interactions}  
  %\VignetteEncoding{UTF-8}   
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

TODO(wew,jg): Can you please add a one-liner of the purpose of the vignette?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
evalAll = TRUE
```

```{r}
a <- c(a = 3, b = 4.5, c = 6, d = 7.5, e = 9)

F1 <- list(L1 = 3, L2 = 4.5)
F2 <- list(L1 = 0, L2 = 3) 


a <- F1$L1 + F2$L1 # 3
b <- F1$L2 + F2$L1 # 4.5
c <- F1$L1 + F2$L2 # 6
d <- F1$L2 + F2$L2 # 7.5

#(F1L1 - F1L2) gv F2L1 = log2(3/4.5) = -0.585 
#(F1L1 - F1L2) gv F2L2 = log2(6/7.5) = -0.321 
    
    
c(a, b, c, d)
```

First, we load the data and do the configuration.

```{r readAnnotation}
datadir <- file.path(find.package("prolfquadata") , "quantdata")

inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")

annotation <- readxl::read_xlsx(inputAnnotation)
annotation <- annotation |> dplyr::filter(sample != "e")
annotation <- annotation |> 
  dplyr::mutate(F1 = dplyr::case_when(sample %in% c("a","c") ~ "L1", TRUE ~ "L2"),
    F2 = dplyr::case_when(sample %in% c("a","b") ~ "L1", TRUE ~ "L2")) |>
  dplyr::arrange(sample)
```


```{r fromPeptideTXT , eval = evalAll}
datadir <- file.path(find.package("prolfquadata") , "quantdata")
inputMQfile <-  file.path(datadir,
                          "MAXQuant_IonStar2018_PXD003881.zip")

data <- prolfquapp::tidyMQ_Peptides(inputMQfile)
length(unique(data$proteins))
```

Read the sample annotation. The sample annotation must contain the `raw.file` name and the explanatory variables of your experiment, e.g. treatment, time point, genetic background, or other factors which you would like to check for confounding.


Then you need to _tell_ `prolfqua` which columns in the data frame contain what
information. You do it using the `AnalysisTableAnnotation` class.


The `AnalysisTableAnnotation` has the following fields that need to be populated:
- fileName
- hierarchy
- factors
- workingIntensity
, and we will discuss it in more detail below.

The `fileName` is the column with the raw file names, however for labelled TMT experiments, it can be used to hold the name of the TMT channel.


The `hierarchy` field describes the structure of the MS data e.g, 
- protein 
- peptides
- modified peptides
- precursor
In case of the MaxQuant `peptide.txt` file we have data on protein level.

In addition, we need to describe the `factors` of the analysis, i.e., the column
containing the explanatory variables. 


```{r}

config <- prolfqua::create_config_MQ_peptide()

res <- prolfquapp::add_annotation(
  data,
  annotation,
  fileName = "raw.file"
)

config$table$factors[["F1."]] = "F1"
config$table$factors[["F2."]] = "F2"
config$table$factorDepth <- 2

data <- prolfqua::setup_analysis(res, config)


lfqdata <- prolfqua::LFQData$new(data, config)
```

Filter the data for small intensities (maxquant reports missing values as 0) and for two peptides per protein.


```{r}
lfqdata$data <- lfqdata$data |> dplyr::filter(!grepl("^REV__|^CON__", protein_Id)) 
lfqdata$filter_proteins_by_peptide_count()
lfqdata$remove_small_intensities()
lfqdata$hierarchy_counts()
```

```{r}
tr <- lfqdata$get_Transformer()
subset_h <- lfqdata$get_copy()
subset_h$data <- subset_h$data |> dplyr::filter(grepl("HUMAN", protein_Id))
subset_h <- subset_h$get_Transformer()$log2()$lfq
lfqdataNormalized <- tr$log2()$robscale_subset(lfqsubset = subset_h)$lfq

lfqAggMedpol <- lfqdataNormalized$get_Aggregator()
lfqAggMedpol$medpolish()
lfqTrans <- lfqAggMedpol$lfq_agg

```


# Model Fitting

defines the contrasts

```{r specifyModel}
lfqTrans$rename_response("abundance")
formula_2_Factors <- prolfqua::strategy_lm("abundance ~ F1. * F2. ")

# specify model definition
modelName  <- "Model"


Contrasts <- c("F1.L1_vs_F1.L2" = "F1.L1 - F1.L2",
               "F2.L1_vs_F2.L2" = "F2.L1 - F2.L2",
               "F1L1_vs_F1L2_gv_F2L1" = "`F1.L1:F2.L1` - `F1.L2:F2.L1`",
               "F1L1_vs_F1L2_gv_F2L2" = "`F1.L1:F2.L2` - `F1.L2:F2.L2`",
               "doFCinF2L1_differfromF2L2" = "`F1L1_vs_F1L2_gv_F2L1` - `F1L1_vs_F1L2_gv_F2L2`"
               )
```

The following line fits the model.

```{r buildModel}
mod <- prolfqua::build_model(lfqTrans, formula_2_Factors)
```


```{r anovaPvaluePlots, fig.cap="p-value distributions for ANOVA analysis."}
mod$anova_histogram(what = "FDR")
```

## ANOVA

Examine proteins with a significant interaction between the two factors treatment and batch.

```{r anovaAnalysis}
ANOVA <- mod$get_anova()
ANOVA |>
  dplyr::filter(factor == "F1.:F2.") |>
  dplyr::arrange(FDR) |>
  head(5)

ANOVA$factor |> unique()
protIntSig <- ANOVA |> dplyr::filter(factor == "F1.") |>
  dplyr::filter(FDR < 0.1)

protInt <-  lfqTrans$get_copy()
protInt$data <- protInt$data[protInt$data$protein_Id %in%
                               protIntSig$protein_Id, ]
protInt$hierarchy_counts()
mean(grepl("ECOLI", protInt$data$protein_Id))

```


```{r fig.with=15, fig.height=15, fig.cap="Proteins with FDR < 0.5 for condition batch interaction in ANOVA."}
#ggpubr::ggarrange(plotlist = protInt$get_Plotter()$boxplots()$boxplot)
```

# Compute contrasts

```{r computeModeratedContrasts}
contr <- prolfqua::ContrastsModerated$new(prolfqua::Contrasts$new(mod, Contrasts))
#contr$get_contrasts_sides()
contrdf <- contr$get_contrasts()
```


The code snippets graph the volcano and ma plot.
```{r graphVolcano}
plotter <- contr$get_Plotter()
plotter$volcano()
plotter$ma_plot()
```

## Annalyse contrasts with missing data imputation

```{r}
lfqTrans$config$table$factorDepth <- 2
# ContrastsMissing$debug("get_contrasts")
contrSimple <- prolfqua::ContrastsMissing$new(lfqdata = lfqTrans,
                                                   Contrasts)
contrdfSimple <- contrSimple$get_contrasts()
# na.omit(contrdfSimple)
pl <- contrSimple$get_Plotter()
pl$histogram_diff()
pl$volcano()
```

## Merge nonimputed and imputed data.



```{r mergeNonImpyted}
dim(contr$get_contrasts())
dim(contrSimple$get_contrasts())

mergedContrasts <- prolfqua::merge_contrasts_results(prefer = contr, add = contrSimple)$merged
cM <- mergedContrasts$get_Plotter()
```

```{r plotNergeNonImpyted}
plot <- cM$volcano()
plot$FDR
```

The `prolfqua` package is described in [@Wolski2022.06.07.494524].

# Session Info

```{r}
sessionInfo()
```

# References






