---
title: "Pipeline Demo: Fit → Write → Score → Aggregate"
author: "Witold Wolski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      toc_collapsed: false
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Pipeline Demo: Fit-Write-Score-Aggregate}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

This vignette demonstrates the **pipeline architecture** for benchmarking.
Instead of fitting models and benchmarking in one monolithic script, the workflow
is split into three decoupled steps:

1. **Fit** — fit models and compute contrasts, write standardized output files
2. **Score** — load contrast result files, create Benchmark objects, compute metrics
3. **Aggregate** — collect benchmark summaries across methods for comparison

Each fitting step produces a directory with `contrasts.tsv` + `metadata.yaml`.
This enables a future Snakemake pipeline: `for each dataset × method → fit → score → aggregate`.

```{r setup, include=FALSE}
library(prolfquabenchmark)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  knitr.table.format = "pipe"
)
```

```{r loadRequiredPackages}
conflicted::conflict_prefer("filter", "dplyr")
```

# Shared Data Preparation

This section is identical to `BenchmarkingIonstarData.Rmd` — load and normalize
the IonStar dataset, aggregate peptides to proteins via median polish.

```{r loadAndPrepareData}
datadir <- file.path(find.package("prolfquadata"), "quantdata")
inputMQfile <- file.path(datadir, "MAXQuant_IonStar2018_PXD003881.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")

mqdata <- list()
mqdata$data <- prolfquapp::tidyMQ_Peptides(inputMQfile)
mqdata$config <- prolfqua::create_config_MQ_peptide()

annotation <- readxl::read_xlsx(inputAnnotation)
res <- dplyr::inner_join(mqdata$data, annotation, by = "raw.file")

mqdata$config$table$factors[["dilution."]] <- "sample"
mqdata$config$table$factors[["run_Id"]] <- "run_ID"
mqdata$config$table$factorDepth <- 1
mqdata$data <- prolfqua::setup_analysis(res, mqdata$config)

lfqdata <- prolfqua::LFQData$new(mqdata$data, mqdata$config)
lfqdata$data <- lfqdata$data |> dplyr::filter(!grepl("^REV__|^CON__", protein_Id))
lfqdata$filter_proteins_by_peptide_count()
lfqdata$remove_small_intensities()

# Normalize using HUMAN subset
tr <- lfqdata$get_Transformer()
subset_h <- lfqdata$get_copy()
subset_h$data <- subset_h$data |> dplyr::filter(grepl("HUMAN", protein_Id))
subset_h <- subset_h$get_Transformer()$log2()$lfq
lfqdataNormalized <- tr$log2()$robscale_subset(lfqsubset = subset_h, preserveMean = FALSE)$lfq

# Median polish aggregation
lfqAggMedpol <- lfqdataNormalized$get_Aggregator()
lfqAggMedpol$medpolish()
protLFQ <- lfqAggMedpol$lfq_agg
```

```{r defineContrasts}
relevantContrasts <- c(
  "dilution_(9/7.5)_1.2"    = "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25"   = "dilution.d - dilution.c",
  "dilution_(6/4.5)_1.3(3)" = "dilution.c - dilution.b",
  "dilution_(4.5/3)_1.5"    = "dilution.b - dilution.a"
)
```

```{r createOutputDir}
output_base <- tempfile("benchmark_pipeline_")
dir.create(output_base)
message("Pipeline output directory: ", output_base)
```

# Step 1: Fit & Write

Each method fits models, computes contrasts, and writes the results using
`write_contrast_results()`. This produces a `contrasts.tsv` + `metadata.yaml`
per method.

## Method: Linear model (lm)

```{r fitLM}
lmmodel <- paste0(protLFQ$config$table$get_response(), " ~ dilution.")
modelFunction <- prolfqua::strategy_lm(lmmodel, model_name = "Model")
modLinearProt <- prolfqua::build_model(protLFQ$data, modelFunction)

contrProt <- prolfqua::Contrasts$new(modLinearProt, relevantContrasts)
contrast_data_lm <- contrProt$get_contrasts()

```

```{r writeLM}
write_contrast_results(
  data = contrast_data_lm,
  path = file.path(output_base, "ionstar_lm"),
  metadata = list(
    dataset = "ionstar_maxquant",
    method = "prolfqua_lm",
    method_description = "Linear model on median-polished proteins",
    input_level = "peptide",
    aggregation = "median_polish",
    normalization = "robscale",
    software_version = paste0("prolfqua ", packageVersion("prolfqua")),
    date = as.character(Sys.Date()),
    contrasts = names(relevantContrasts),
    ground_truth = list(
      id_column = "protein_id",
      positive = list(label = "ECOLI", pattern = "ECOLI"),
      negative = list(label = "HUMAN", pattern = "HUMAN")
    )
  )
)
```

## Method: Moderated linear model (lm + moderation)

```{r fitLMmod}
contrProtModerated <- prolfqua::ContrastsModerated$new(contrProt)
contrast_data_lm_mod <- contrProtModerated$get_contrasts()
```

```{r writeLMmod}
write_contrast_results(
  data = contrast_data_lm_mod,
  path = file.path(output_base, "ionstar_lm_mod"),
  metadata = list(
    dataset = "ionstar_maxquant",
    method = "prolfqua_lm_mod",
    method_description = "Linear model with variance moderation on median-polished proteins",
    input_level = "peptide",
    aggregation = "median_polish",
    normalization = "robscale",
    software_version = paste0("prolfqua ", packageVersion("prolfqua")),
    date = as.character(Sys.Date()),
    contrasts = names(relevantContrasts),
    ground_truth = list(
      id_column = "protein_id",
      positive = list(label = "ECOLI", pattern = "ECOLI"),
      negative = list(label = "HUMAN", pattern = "HUMAN")
    )
  )
)
```

## Method: Imputation (missing value model)

```{r fitImputed}
contrImp <- prolfqua::ContrastsMissing$new(protLFQ, relevantContrasts)
contrast_data_imp <- contrImp$get_contrasts()
```

```{r writeImputed}
write_contrast_results(
  data = contrast_data_imp,
  path = file.path(output_base, "ionstar_missing"),
  metadata = list(
    dataset = "ionstar_maxquant",
    method = "prolfqua_missing",
    method_description = "Median polish and missingness modelling",
    input_level = "peptide",
    aggregation = "median_polish",
    normalization = "robscale",
    software_version = paste0("prolfqua ", packageVersion("prolfqua")),
    date = as.character(Sys.Date()),
    contrasts = names(relevantContrasts),
    ground_truth = list(
      id_column = "protein_id",
      positive = list(label = "ECOLI", pattern = "ECOLI"),
      negative = list(label = "HUMAN", pattern = "HUMAN")
    )
  )
)
```

## Method: Merged (lm + imputation, moderated)

```{r fitMerged}
all_merged <- prolfqua::merge_contrasts_results(prefer = contrProt, add = contrImp)
merged_mod <- prolfqua::ContrastsModerated$new(all_merged$merged)
contrast_data_merged <- merged_mod$get_contrasts()
```

```{r writeMerged}
write_contrast_results(
  data = contrast_data_merged,
  path = file.path(output_base, "ionstar_merged"),
  metadata = list(
    dataset = "ionstar_maxquant",
    method = "prolfqua_merged",
    method_description = "Merge of lm moderated and imputed contrasts",
    input_level = "peptide",
    aggregation = "median_polish",
    normalization = "robscale",
    software_version = paste0("prolfqua ", packageVersion("prolfqua")),
    date = as.character(Sys.Date()),
    contrasts = names(relevantContrasts),
    ground_truth = list(
      id_column = "protein_id",
      positive = list(label = "ECOLI", pattern = "ECOLI"),
      negative = list(label = "HUMAN", pattern = "HUMAN")
    )
  )
)
```

### Inspect output files

```{r inspectOutput}
list.files(output_base, recursive = TRUE)
```

```{r showTSV}
lm_data <- utils::read.delim(
  file.path(output_base, "ionstar_lm", "contrasts.tsv"),
  nrows = 5
)
knitr::kable(lm_data, caption = "First 5 rows of ionstar_lm/contrasts.tsv")
```

```{r showYAML}
cat(readLines(file.path(output_base, "ionstar_lm", "metadata.yaml")), sep = "\n")
```

# Step 2: Score

Load each contrast result directory with `benchmark_from_file()` and compute
benchmark metrics. No knowledge of the fitting step is needed — the ground
truth specification comes from `metadata.yaml`.

```{r scoreMethods}
method_dirs <- list.dirs(output_base, recursive = FALSE)
names(method_dirs) <- basename(method_dirs)

benchmarks <- lapply(method_dirs, benchmark_from_file)
```

```{r showROC, fig.width=8, fig.height=6}
benchmarks$ionstar_lm$plot_ROC(xlim = 0.15)
```

```{r showPrecisionRecall}
benchmarks$ionstar_lm_mod$plot_precision_recall()
```

```{r showFDR}
benchmarks$ionstar_merged$plot_FDRvsFDP()
```

# Step 3: Aggregate & Compare

Use `to_summary_table()` to extract metrics from each Benchmark, then combine
with `collect_benchmark_results()` for cross-method comparison.

```{r extractSummaries}
summary_files <- character()
for (method_name in names(benchmarks)) {
  bench <- benchmarks[[method_name]]
  bench$complete(FALSE)
  summary <- bench$to_summary_table(dataset = "ionstar_maxquant")

  outfile <- file.path(output_base, method_name, "benchmark_results.tsv")
  write_benchmark_results(summary, outfile)
  summary_files <- c(summary_files, outfile)
}
```

```{r collectResults}
combined <- collect_benchmark_results(summary_files)
knitr::kable(
  combined |> dplyr::filter(contrast == "all"),
  caption = "Benchmark summary across methods (all contrasts pooled)",
  digits = 2
)
```

```{r comparePlot, fig.width=9, fig.height=6}
combined_all <- combined |> dplyr::filter(contrast == "all")
long <- tidyr::pivot_longer(
  combined_all,
  cols = c("AUC", "pAUC_10", "pAUC_20", "AP", "pAP_50", "pAP_80"),
  names_to = "metric",
  values_to = "value"
)
ggplot2::ggplot(
  long,
  ggplot2::aes(x = value, y = n_total, color = model_name, shape = score)
) +
  ggplot2::geom_point(size = 3) +
  ggplot2::facet_wrap(~metric, scales = "free_x") +
  ggplot2::labs(
    x = "Metric score (%)", y = "Nr. quantified proteins",
    color = "Method", shape = "Ranking score"
  )
```

```{r summaryMetrics, fig.width=10, fig.height=6}
# Geometric mean across contrasts — complements the pooled "all" row above
summary_list <- lapply(benchmarks, function(b) {
  b$complete(FALSE)
  b$summary_metrics()
})
combined_summary <- dplyr::bind_rows(summary_list)

long_summary <- tidyr::pivot_longer(
  combined_summary,
  cols = c("AUC", "pAUC_10", "pAUC_20", "AP", "pAP_50", "pAP_80"),
  names_to = "metric",
  values_to = "value"
)
ggplot2::ggplot(
  long_summary,
  ggplot2::aes(x = value, y = n_total, color = model_name, shape = score)
) +
  ggplot2::geom_point(size = 3) +
  ggplot2::facet_wrap(~metric, scales = "free_x") +
  ggplot2::labs(
    x = "Geometric mean of metric across contrasts (%)",
    y = "Mean nr. quantified proteins",
    color = "Method", shape = "Ranking score",
    title = "Cross-contrast summary (geometric mean)"
  )
```

# Validation: Round-Trip Comparison

Verify that the file-based pipeline produces the same AUC values as the
direct (in-memory) path used in the original vignette.

```{r validateRoundTrip}
# Direct path (as in BenchmarkingIonstarData.Rmd)
ttd_direct <- ionstar_bench_preprocess(contrast_data_lm)
bench_direct <- make_benchmark(
  ttd_direct$data,
  model_description = "med. polish and lm",
  model_name = "prolfqua_lm"
)
pauc_direct <- bench_direct$pAUC()

# File-based path
pauc_file <- benchmarks$ionstar_lm$pAUC()

# Compare — AUC and AP values should be identical
comparison <- dplyr::inner_join(
  pauc_direct |> dplyr::select(contrast, what, AUC_direct = AUC,
                                pAUC_10_direct = pAUC_10,
                                AP_direct = AP),
  pauc_file |> dplyr::select(contrast, what, AUC_file = AUC,
                              pAUC_10_file = pAUC_10,
                              AP_file = AP),
  by = c("contrast", "what")
)
comparison$AUC_match <- abs(comparison$AUC_direct - comparison$AUC_file) < 0.01
comparison$pAUC_10_match <- abs(comparison$pAUC_10_direct - comparison$pAUC_10_file) < 0.01
comparison$AP_match <- abs(comparison$AP_direct - comparison$AP_file) < 0.01

knitr::kable(comparison, caption = "Round-trip validation: direct vs file-based AUC and AP", digits = 3)

stopifnot(all(comparison$AUC_match), all(comparison$pAUC_10_match), all(comparison$AP_match))
message("Round-trip validation PASSED: all AUC and AP values match.")
```

# Summary

The pipeline architecture decouples fitting from scoring:

- **`write_contrast_results()`** serializes contrast results + metadata to `contrasts.tsv` + `metadata.yaml`
- **`benchmark_from_file()`** loads results, applies ground truth annotation from metadata, creates a `Benchmark` object
- **`to_summary_table()`** extracts AUC metrics into a flat table
- **`collect_benchmark_results()`** combines summaries across methods

This enables:

- Independent execution of fitting steps (parallelizable via Snakemake)
- Reproducible scoring without re-running expensive model fitting
- Easy addition of new methods (just produce `contrasts.tsv` + `metadata.yaml`)
- Cross-method comparison from summary TSV files

```{r cleanup, include=FALSE}
unlink(output_base, recursive = TRUE)
```
