---
title: "DEA benchmark CPTAC/MaxQuant/peptide.txt (prolfqua)"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      toc_collapsed: false
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{DEA benchmark CPTAC/MaxQuant/peptide.txt (prolfqua)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence

---

```{r setup, include=FALSE}
library(prolfquabenchmark)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
evalAll <- require("msdata")
```

# Introduction

The r-package $msdata$ provides a subset of the data of the 6th study of the Clinical Proteomic Technology Assessment for Cancer (CPTAC). In this experiment, the Sigma Universal Protein Standard mixture 1 (UPS1) containing $48$ different human proteins was spiked in a protein background of 60 ng/$\mu L$ Saccharomyces cerevisiae strain BY4741. Two different spike-in concentrations were used, 6A ($0.25$ fmol UPS1 proteins/$\mu L$) and 6B ($0.74$ fmol UPS1 proteins/$\mu L$). Three replicates are available for each concentration.


```{r cars, eval = evalAll}
library(dplyr)
library(tidyr)
library(ggplot2)
tmp <- msdata::quant(full.names = TRUE)
xx <- read.csv(tmp, sep = "\t")
peptides <- prolfquapp::tidyMQ_Peptides(xx)

annotation <- data.frame(raw.file = unique(peptides$raw.file))
annotation <- annotation |> dplyr::mutate(group = gsub("^6","",raw.file)) |> mutate(group = gsub("_[0-9]$", "", group))

peptides <- dplyr::inner_join(annotation, peptides)


config <- prolfqua::create_config_MQ_peptide()
config$table$factors["group."] = "group"

peptides <- prolfqua::setup_analysis(peptides, config)
lfqpeptides <- prolfqua::LFQData$new(peptides, config)
lfqpeptides$filter_proteins_by_peptide_count()
lfqpeptides$remove_small_intensities()

lfqpeptides$hierarchy_counts()

lfqpeptides$factors()
tmp <- lfqpeptides$data |> filter(grepl("_UPS", protein_Id))

```

# Transform and Aggregate

First we transform the data using vsn::justvsn function and afterwards we estimate protein intensities using
Tukeys median polish.

```{r, eval = evalAll}
tr <- lfqpeptides$get_Transformer()
tr <- tr$intensity_matrix(vsn::justvsn)
lfqtransformed <- tr$lfq
agg <- lfqtransformed$get_Aggregator()
agg$medpolish()
pl <- agg$plot()
```

```{r aggregationResults, fig.cap = "Peptide abundances and protein abundance estimates.", eval = evalAll}
gridExtra::grid.arrange(grobs = pl$plots[33:38])
lfqProt <- agg$lfq_agg

```

# Fit model with prolfqua

```{r, eval = evalAll}
library(prolfqua)

model <- paste(lfqProt$response(), " ~ group. ")
lm <- prolfqua::strategy_lm(model)
lmmodl <- prolfqua::build_model(lfqProt, lm)
contrast <- c("b_vs_a" = "group.b - group.a")

clm <- prolfqua::Contrasts$new(lmmodl, contrast)
tt <- clm$get_contrasts()

csimp <- prolfqua::ContrastsMissing$new(lfqProt, contrast)
tt <- csimp$get_contrasts()

merge <- prolfqua::merge_contrasts_results(clm, csimp)
cmod <- ContrastsModerated$new(merge$merge)

pl <- cmod$get_Plotter()
pl$volcano()$FDR

```

Setting up benchmark. UPS proteins are true positives, while YEAST proteins are true negatives.
Contaminants are removed. This leaves 1460 proteins.

```{r, eval = evalAll}
ttd <- prolfquabenchmark::cptac_bench_preprocess(cmod$get_contrasts(), idcol = "protein_Id")
benchmark_merged <- make_benchmark(
  ttd$data,
  model_description = "merge of prot moderated and imputed",
  model_name = "prolfqua_merged")


sumarry <- benchmark_merged$smc$summary
knitr::kable(sumarry, caption = "nr of proteins with 0, 1, 2, 3 missing contrasts.",format = "html")

```

```{r, eval = evalAll}
res <- benchmark_merged$pAUC_summaries()
knitr::kable(res$ftable$content,caption = res$ftable$caption)
res$barp
```

```{r writeContrasts, eval = evalAll}
write_contrast_results(
  cmod$get_contrasts(),
  path = "../inst/Benchresults/prolfqua_CPTAC",
  metadata = list(
    dataset = "CPTAC",
    method = "prolfqua",
    method_description = "prolfqua merged + moderated on CPTAC MQ peptides + VSN",
    input_file = "msdata::quant()",
    software_version = paste0("prolfqua ", packageVersion("prolfqua")),
    date = as.character(Sys.Date()),
    ground_truth = list(
      id_column = "protein_id",
      positive = list(label = "UPS", pattern = "_UPS"),
      negative = list(label = "YEAST", pattern = "_YEAST")
    )
  ))
```
