---
title: "DEA benchmark IonStar/FragPipeV14/MSstats.csv (prolfqua, msqrob2)"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{DEA benchmark IonStar/FragPipeV14/MSstats.csv (prolfqua, msqrob2)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

Please download and install the `prolfquadata` package from github

```{r setup, include=FALSE}
library(prolfquabenchmark)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 10
)
evalAll <- requireNamespace("msqrob2", quietly = TRUE)
```

```{r conflicted, eval = evalAll}
conflicted::conflict_prefer("filter", "dplyr")

```


# Data preprocessing



## Load data


```{r loaddata, eval = evalAll}
datadir <- file.path(find.package("prolfquadata") , "quantdata")
inputFragfile <-  file.path(datadir, "MSFragger_IonStar2018_PXD003881.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")
annotation <- readxl::read_xlsx(inputAnnotation)

unzip(inputFragfile, list = TRUE)
msstats <- read.csv(
    unz(inputFragfile,"IonstarWithMSFragger/MSstats.csv"),
    header = TRUE,
    sep = ",",
    stringsAsFactors = FALSE)



msstats <- tibble::tibble(msstats)
msstats$Run <- tolower(msstats$Run)

merged <- dplyr::inner_join(annotation, msstats, by = c("raw.file" = "Run"))

summedPep <- merged |> dplyr::group_by(raw.file, run_ID, sample, ProteinName, PeptideSequence) |> dplyr::summarize(nprecursor = dplyr::n(),pepIntensity = sum(Intensity, na.rm = TRUE), .groups = "drop")


```

## Create prolfqua configuration

```{r setupConfigs, eval = evalAll}
atable <- prolfqua::AnalysisConfiguration$new()
atable$fileName = "raw.file"
atable$hierarchy[["protein_Id"]] <- c("ProteinName")
atable$hierarchy[["peptide_Id"]] <- c("PeptideSequence")

atable$hierarchyDepth <- 1
atable$set_response("pepIntensity")
atable$factors[["dilution."]] = "sample"
atable$factors[["run"]] = "run_ID"
atable$factorDepth <- 1

config <- prolfqua::AnalysisConfiguration$new(atable)

adata <- prolfqua::setup_analysis(summedPep, config)
lfqdata <- prolfqua::LFQData$new(adata, config)
lfqdata$hierarchy_counts()
lfqdata$filter_proteins_by_peptide_count()
lfqdata$hierarchy_counts()
lfqdata$remove_small_intensities()
lfqdata$hierarchy_counts()

knitr::kable(lfqdata$hierarchy_counts(), caption = "number of proteins and peptides.")

```

## Normalize data using human proteins

```{r createPlotter, eval = evalAll}
pl <- lfqdata$get_Plotter()
pl$intensity_distribution_density()
subset_h <- lfqdata$get_copy()$get_Transformer()$log2()$lfq
subset_h$data <- subset_h$data |> dplyr::filter(grepl("HUMAN", protein_Id))
tr <- lfqdata$get_Transformer()
lfqdataPeptideNorm <- tr$log2()$robscale_subset(lfqsubset = subset_h)$lfq
pl <- lfqdataPeptideNorm$get_Plotter()
pl$intensity_distribution_density()
hm <- pl$NA_heatmap()
```

```{r showHeatmap, eval = evalAll}
hm
```

## Aggregate proteins

```{r aggregateProteins, eval = evalAll}
agg <- lfqdataPeptideNorm$get_Aggregator()
agg$medpolish()
lfqdataNormalized <- agg$lfq_agg
lfqdataNormalized$hierarchy_counts()
```


# DEA using prolfqua

```{r specifyContrasts, eval = evalAll}
Contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c",
  "dilution_(6/4.5)_1.3(3)" =   "dilution.c - dilution.b",
  "dilution_(4.5/3)_1.5" =   "dilution.b - dilution.a"
)


lmmodel <- "~ dilution."
lmmodel <- paste0(lfqdataNormalized$config$table$get_response() , lmmodel)

modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "Model")


mod <- prolfqua::build_model(lfqdataNormalized$data, modelFunction)
contr <- prolfqua::Contrasts$new(mod, Contrasts)
contrimp <- prolfqua::ContrastsMissing$new(lfqdataNormalized, Contrasts)

merged <- prolfqua::merge_contrasts_results(contr, contrimp)
mergedmod <- prolfqua::ContrastsModerated$new(merged$merged)

cp <- mergedmod$get_Plotter()
cp$volcano()$FDR

```


## Benchmarking


```{r benchmark, eval = evalAll}
ttd <- ionstar_bench_preprocess(mergedmod$get_contrasts())
benchmark_prolfqua <- make_benchmark(ttd$data,
                                             model_description = "prolfqua_merged",
                                             model_name = "prolfqua_merged",
                                             FDRvsFDP = list(list(score = "FDR", desc = FALSE))
)


knitr::kable(benchmark_prolfqua$pAUC_summaries()$ftable$content)


prolfqua::table_facade(benchmark_prolfqua$smc$summary, "Nr of estimated contrasts")
benchmark_prolfqua$plot_score_distribution()
benchmark_prolfqua$plot_ROC(0.05)
benchmark_prolfqua$plot_FDRvsFDP()

```


# DEA using msqrob2

```{r aggregateMedianPolish, eval=TRUE, include = FALSE, eval = evalAll}
library(QFeatures)
library(msqrob2)

se <- prolfqua::LFQDataToSummarizedExperiment(lfqdata = lfqdataPeptideNorm)
pe <- QFeatures::QFeatures(list(peptide = se), colData = colData(se))

my_medianPolish <- function(x, verbose = FALSE, ...){
  medpol <- stats::medpolish(x, na.rm = TRUE, trace.iter = verbose, maxiter = 10)
  return(medpol$overall + medpol$col)
}

pe <- QFeatures::aggregateFeatures(
  pe,
  i = "peptide", fcol = "protein_Id",
  name = "protein", fun = my_medianPolish,
)
```

## Defining Contrasts and computing group comparisons

As usual, two steps are required, first fit the models, then compute the contrasts.

```{r runMSRob,  include=TRUE, eval = evalAll}
prlm <- msqrobHurdle(pe,
                     i = "protein",
                     formula = ~dilution.,
                     overwrite = TRUE)


```

Since msqrob does not report average abundances, we are computing them for each contrast.

```{r generateAverageAbundances, eval = evalAll}
st <- lfqdataNormalized$get_Stats()
protAbundanceIngroup <- st$stats()
protAbundanceIngroup <- protAbundanceIngroup |>
  tidyr::pivot_wider(
    id_cols = protein_Id,
    names_from = dilution.,
    names_prefix = "abd.",
    values_from = meanAbundance)


```


```{r computeHurdleContrasts, eval = evalAll}
L <- makeContrast(c("dilution.e-dilution.d=0",
                    "dilution.d-dilution.c=0",
                    "dilution.c-dilution.b=0",
                    "dilution.b=0"),
                  parameterNames = c("dilution.e",
                                     "dilution.d",
                                     "dilution.c",
                                     "dilution.b"))
prlm <- hypothesisTestHurdle(prlm, i = "protein", L, overwrite = TRUE)
protAbundanceIngroup <- protAbundanceIngroup |> dplyr::mutate( avgAbd.e.d =  mean( c(abd.e,abd.d), na.rm = TRUE)  )
protAbundanceIngroup <- protAbundanceIngroup |> dplyr::mutate( avgAbd.d.c =  mean( c(abd.d,abd.c), na.rm = TRUE)  )
protAbundanceIngroup <- protAbundanceIngroup |> dplyr::mutate( avgAbd.c.b =  mean( c(abd.c,abd.b), na.rm = TRUE)  )
protAbundanceIngroup <- protAbundanceIngroup |> dplyr::mutate( avgAbd.b.a =  mean( c(abd.b,abd.a), na.rm = TRUE)  )


xx <- rowData(prlm[["protein"]])
hurdle <- xx[grepl("hurdle_",names(xx))]

res <- list()
for (i in names(hurdle)) {
  hurdle[[i]]$contrast <- i
  res[[i]] <- prolfqua::matrix_to_tibble(hurdle[[i]], preserve_row_names = "name")
}

hurdle <- dplyr::bind_rows(res)

```


Now we need to merge the results of both models msqrobHurdleIntensity and msqrobHurdleCount. To find out which models were not estimated by `msqrobHurdleIntensity` we check for NA's and use the `anti_join` to select those from the `msqrobHurdleCount` models.

```{r merge, eval = evalAll}
logFC <- hurdle |> dplyr::select("name","contrast", starts_with("logFC"))
logFC <- dplyr::filter(logFC ,!is.na(logFCt))

logFC$modelName <- "msqrobHurdleIntensity"
names(logFC) <- c("name","contrast","logFC","se","df","t","pval","modelName")

logOR <- hurdle |> dplyr::select("name","contrast", starts_with("logOR"))
logOR$modelName <- "msqrobHurdleCount"
names(logOR) <- c("name","contrast","logFC","se","df","t","pval","modelName")


ddd <- dplyr::anti_join(logOR , logFC, by = c("name", "contrast"))
all <- dplyr::bind_rows(ddd , logFC) |> dplyr::arrange(contrast, name)
all <- prolfqua::adjust_p_values(all, column = "pval", group_by_col = "contrast")
all$contrast |> unique()

protAbundanceIngroup <- protAbundanceIngroup |>
  dplyr::select(-starts_with("abd")) |>
  tidyr::pivot_longer(starts_with("avgAbd"), names_to = "contrast" ,values_to = "avgAbd")

protAbundanceIngroup <- protAbundanceIngroup |>
  dplyr::mutate(contrast =
           dplyr::case_when(contrast == "avgAbd.e.d" ~ "dilution_(9/7.5)_1.2",
                     contrast == "avgAbd.d.c" ~ "dilution_(7.5/6)_1.25",
                     contrast == "avgAbd.c.b" ~ "dilution_(6/4.5)_1.3(3)",
                     contrast == "avgAbd.b.a" ~ "dilution_(4.5/3)_1.5",
                     TRUE ~ "something wrong"))
all <- all |> dplyr::mutate(contrast = dplyr::case_when(
  contrast == "hurdle_dilution.e - dilution.d" ~ "dilution_(9/7.5)_1.2",
  contrast == "hurdle_dilution.d - dilution.c" ~ "dilution_(7.5/6)_1.25",
  contrast == "hurdle_dilution.c - dilution.b" ~ "dilution_(6/4.5)_1.3(3)",
  contrast == "hurdle_dilution.b" ~ "dilution_(4.5/3)_1.5",
  TRUE ~ "something wrong"))
stopifnot(sum(all$contrast == "something wrong") == 0 )
stopifnot(sum(all$contrast == "something wrong") == 0 )


bb <- dplyr::inner_join(all, protAbundanceIngroup, by = c("name" = "protein_Id", "contrast" = "contrast"))

```


## Benchmarking

Here we use `proflqua` benchmark functions to generate some summaries.

```{r setUPBenchmark, eval = evalAll}

ttd <- ionstar_bench_preprocess( bb , idcol = "name" )
benchmark_msqrob <- make_benchmark(ttd$data,
                                             contrast = "contrast",
                                             toscale = c("pval"),
                                             fcestimate = "logFC",
                                             benchmark = list(
                                               list(score = "logFC", desc = TRUE),
                                               list(score = "t", desc = TRUE),
                                               list(score = "scaled.pval", desc = TRUE)
                                             ),
                                             model_description = "msqrob_QFeature",
                                             model_name = "msqrob2",
                                             FDRvsFDP = list(list(score = "FDR", desc = FALSE))
                                             , hierarchy = c("name"), summarizeNA = "t"
)

sumarry <- benchmark_msqrob$smc$summary
prolfqua::table_facade(sumarry, caption = "nr of proteins with 0, 1, 2, 3 missing contrasts.")

```


```{r prepBenchmarkforComparison, include= FALSE, eval = evalAll}
xdd <- ttd$data |> dplyr::rename(protein_Id = name ,
                                 contrast = contrast,
                                 avgInt = avgAbd,
                                 diff = logFC,
                                 statistic = t,
                                 p.value = pval,
                                 FDR = FDR
)

#prolfqua::Benchmark$undebug("get_confusion_FDRvsFDP")
benchmark2_msqrob <- make_benchmark(xdd, model_description = "msqrob2", model_name = "msqrob2")

```




```{r rocCurve, fig.cap="ROC curves", eval = evalAll}
res <- benchmark_msqrob$pAUC_summaries()
knitr::kable(res$ftable$content,caption = res$ftable$caption)
res$barp
```

```{r pAUC02, fig.cap="plot ROC curves", eval = evalAll}
#res$ftable
benchmark_msqrob$plot_ROC(xlim = 0.2)
```

```{r fdrfdp, fig.cap = "plot FDR vs FDP", eval = evalAll}
benchmark_msqrob$plot_FDRvsFDP()
```

```{r fdptpr, eval = evalAll}
benchmark_msqrob$plot_precision_recall()
```

# Comparing DEA results from prolfqua and msqrob2

```{r compareResults, eval = evalAll}
allB <- list(benchmark_prolfqua = benchmark_prolfqua,
             benchmark_msqrob = benchmark_msqrob)

bdir <- file.path("../inst/Benchresults/", format(Sys.Date(), "%Y%m%d"))
if (!dir.exists(bdir)) {dir.create(bdir)}

saveRDS(allB,
  file.path("../inst/Benchresults/", format(Sys.Date(), "%Y%m%d"), "FragPipev14_comb.RDS"))

prolfqua <- allB$benchmark_prolfqua$pAUC_summaries()$ftable$content
prolfqua$package <- "prolfqua"
msqrob2 <- allB$benchmark_msqrob$pAUC_summaries()$ftable$content
msqrob2$package <- "msqrob2"
tmp <- dplyr::bind_rows(prolfqua, msqrob2)

tmp$what[tmp$what == "statistic"] <- "t_statistic"
tmp$what[tmp$what == "scaled.pval"] <- "scaled.p.value"
tmp$what[tmp$what == "scaled.pvalue"] <- "scaled.p.value"

tmp$what[tmp$what == "logFC"] <- "diff"
tmp$what[tmp$what == "log2FC"] <- "diff"
tmp$what[tmp$what == "t"] <- "t_statistic"
tmp$what[tmp$what == "Tvalue"] <- "t_statistic"
tmp$what |> unique()

tmp |> ggplot2::ggplot(ggplot2::aes(x = what, y = pAUC_10, group = package, color = NULL, fill = package)) +
  ggplot2::geom_bar(stat = "identity", position = ggplot2::position_dodge()) +
  ggplot2::facet_wrap(~ contrast) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))

all <- tmp |> dplyr::filter(contrast == "all")

knitr::kable(tmp |> dplyr::filter(contrast == "all"), caption = "Table")

ggplot2::ggplot(all, ggplot2::aes(x = package, y = pAUC_10)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::facet_wrap(~what) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::xlab("")

```


```{r compareEstimatedContrasts, fig.cap= "Number of comparisons for each method", eval = evalAll}

a2 <- allB$benchmark_prolfqua$smc$summary
a2$name <- allB$benchmark_prolfqua$model_name
names(a2)[2] <- "protein_Id"

a3 <- allB$benchmark_msqrob$smc$summary
a3$name <- allB$benchmark_msqrob$model_name
names(a3)[2] <- "protein_Id"

dd <- dplyr::bind_rows(list(a2, a3))
dd <- dd |> dplyr::mutate(nrcontrasts = protein_Id * (4 - nr_missing))
dds <- dd |> dplyr::group_by(name) |> dplyr::summarize(nrcontrasts = sum(nrcontrasts))
dds$percent <- dds$nrcontrasts / max(dds$nrcontrasts) * 100

nrgg <- dds |> ggplot2::ggplot(ggplot2::aes(x = name, y = nrcontrasts)) +
  ggplot2::geom_bar(stat = "identity", fill = "white", colour = "black") +
  ggplot2::coord_cartesian(ylim = c(min(dds$nrcontrasts) - 200, max(dds$nrcontrasts) + 10)) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::geom_text(ggplot2::aes(label = paste0(round(nrcontrasts, digits = 1), paste0("  (", round(percent, digits = 1), "%)"))),
            vjust = 0, hjust = -0.2, angle = -90)
nrgg

```


