---
title: "Benchmarking robust linear model using the Ionstar Dataset"
author: "Witold E Wolski"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      toc_collapsed: false
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Benchmarking robust linear model using the Ionstar Dataset} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

Please download and install the `prolfquadata` package from github

```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo = TRUE,
message = FALSE,
warning = FALSE,
fig.width = 5,
fig.height = 5,
knitr.table.format = "pipe"
)
```

```{r loadRequiredPackages}
conflicted::conflict_prefer("filter", "dplyr")

```

Decide if you work with all data or for speedup with subset of data:

```{r}
SUBSET <- FALSE
SUBSETNORM <- TRUE
SAVE <- TRUE

```

We start by loading the IonStar dataset and the annotation from the `prolfquadata` package.

```{r loadMQData}

datadir <- file.path(find.package("prolfquadata") , "quantdata")
inputMQfile <-  file.path(datadir,
                          "MAXQuant_IonStar2018_PXD003881.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")
mqdata <- list()

mqdata$data <- prolfquapp::tidyMQ_Peptides(inputMQfile)
length(unique(mqdata$data$proteins))

mqdata$config <- prolfqua::create_config_MQ_peptide()


annotation <- readxl::read_xlsx(inputAnnotation)
res <- dplyr::inner_join(
  mqdata$data,
  annotation,
  by = "raw.file"
)

```

The `setup_analysis` asserts that all columns specified in the configruation are present in the data.
For more details about the `prolfqua` configuration see the vignette "Creating Configurations".

```{r annotateData}
mqdata$config$table$factors[["dilution."]] = "sample"
mqdata$config$table$factors[["run_Id"]] = "run_ID"
mqdata$config$table$factorDepth <- 1
mqdata$data <- prolfqua::setup_analysis(res, mqdata$config)

```

## Data normalization

First we remove all contaminant, decoy proteins from the list, than we remove 0 intensity values, then filter for 2 peptides per protein.

```{r filterForHumans}
lfqdata <- prolfqua::LFQData$new(mqdata$data, mqdata$config)
lfqdata$data <- lfqdata$data |> dplyr::filter(!grepl("^REV__|^CON__", protein_Id)) 
sr <- lfqdata$get_Summariser()
lfqdata$remove_small_intensities()
sr <- lfqdata$get_Summariser()
sr$hierarchy_counts()

```

We will normalize the data using the 'LFQTransformer' class.
Since we know that the Human proteins are the Matrix in the experiment we will normalize the data using HUMAN proteins only.
To this task we subset the dataset by filtering for HUMAN proteins only and then use the `LFQDataTransformer` to normalize the data.

```{r}
tr <- lfqdata$get_Transformer()
subset_h <- lfqdata$get_copy()
subset_h$data <- subset_h$data |> dplyr::filter(grepl("HUMAN", protein_Id))
subset_h <- subset_h$get_Transformer()$log2()$lfq
lfqdataNormalized <- tr$log2()$robscale_subset(lfqsubset = subset_h,  preserveMean = FALSE )$lfq

```

The figures below show the intensity distribution before and after normalization.

```{r plotIntensityDistributions}

before <- lfqdata$get_Plotter()
before$intensity_distribution_density()

after <- lfqdataNormalized$get_Plotter()
after$intensity_distribution_density()

```

Create a sample of N proteins to speed up computations of models and contrasts.

```{r createDataSubset}
if (SUBSET) {
  N <- 200
  mqdataSubset <- lfqdata$get_sample(size = N, seed = 2020)
  lfqNormSubset <- lfqdataNormalized$get_sample(size = N, seed = 2020)
  lfqNormSubset$hierarchy_counts()
} else {
  mqdataSubset <- lfqdata$get_copy()  
  lfqNormSubset <- lfqdataNormalized$clone()
  lfqNormSubset$hierarchy_counts()
}
```


## Fitting a robust linear model to peptide abundances

```{r buildModelrlm}
df.residual.rlm <- function(object, ...) {
  return( sum(object$w) - object$rank)
}


sigma.rlm <- function(object, ...) {
  sqrt(sqrt(sum(object$w * object$resid^2) / (sum(object$w) - object$rank)))
}


rlmmodel <- "~ dilution."
rlmmodel <- paste0(lfqNormSubset$config$table$get_response() , rlmmodel)
lfqNormSubset$config$table$hierarchyDepth <- 1

modelFunction <- prolfqua::strategy_rlm( rlmmodel, model_name = "Model")

mod_rlm_ProtLevel <- prolfqua::build_model(lfqNormSubset$data, modelFunction)
mod_rlm_ProtLevel$get_anova()


mod_rlm_ProtLevel$anova_histogram()$plot
```


# Computing Contrasts

Once models are fitted contrasts can be computed.
The R code below defines all possible contrasts among conditions for the ionstar dataset.

```{r defineContrasts}
DEBUG <- FALSE

Contrasts <- c(
  "dilution_(9/3)_3" =   "dilution.e - dilution.a",
  "dilution_(9/4.5)_2" =   "dilution.e - dilution.b",
  "dilution_(9/6)_1.5" =   "dilution.e - dilution.c",
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  
  "dilution_(7.5/3)_2.5" =   "dilution.d - dilution.a",
  "dilution_(7.5/4.5)_1.6(6)" =   "dilution.d - dilution.b",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c",
  
  "dilution_(6/3)_2" =   "dilution.c - dilution.a",
  "dilution_(6/4.5)_1.3(3)" =   "dilution.c - dilution.b",
  
  "dilution_(4.5/3)_1.5" =   "dilution.b - dilution.a"
)


tt <- Reduce(rbind, strsplit(names(Contrasts),split = "_"))
tt <- data.frame(tt)[,2:3]
colnames(tt) <- c("ratio" , "expected fold-change")
tt <- tibble::add_column(tt, contrast =  Contrasts, .before = 1)
prolfqua::table_facade(
  tt,
  caption = "All possible Contrasts given 5 E. coli dilutions of the Ionstar Dataset", digits = 1)

```

```{r usedContrasts}
relevantContrasts <- c("dilution_(9/7.5)_1.2",
                       "dilution_(7.5/6)_1.25",
                       "dilution_(6/4.5)_1.3(3)",
                       "dilution_(4.5/3)_1.5" )

tt <- Reduce(rbind, strsplit(relevantContrasts,split = "_"))
tt <- data.frame(tt)[,2:3]
colnames(tt) <- c("ratio" , "expected fold-change")
tt <- tibble::add_column(tt, contrast =  Contrasts[names(Contrasts) %in% relevantContrasts], .before = 1)
prolfqua::table_facade(tt, caption = "Contrasts used for benchmark.", digits = 1)
relevantContrasts <- Contrasts[names(Contrasts) %in% relevantContrasts]

```


## Contrasts from robust linear model


```{r contrastModel_RLM, include = FALSE}
contrProt_RLM <- prolfqua::Contrasts$new(mod_rlm_ProtLevel, relevantContrasts)
pl <- contrProt_RLM$get_Plotter()
pl$volcano()
pl$histogram()

```


```{r benchmarkModel_RLM , include = FALSE}
ttd <- prolfqua::ionstar_bench_preprocess(contrProt_RLM$get_contrasts())


benchmark_Prot_RLM <- prolfqua::make_benchmark(
    ttd$data,
    model_description = "med. polish and rlm",
    model_name = "prolfqua_rlm"
)

prolfqua::table_facade(benchmark_Prot_RLM$smc$summary, caption = "Nr of proteins with Nr of not estimated contrasts.", digits = 1)

benchmark_Prot_RLM$plot_FDRvsFDP()

```





### Adding Moderation to RLM model

```{r contrastModelRLM1Moderation}

contrProtRLMModerated <- prolfqua::ContrastsModerated$new(contrProt_RLM)
contrProtRLMModerated$get_Plotter()$volcano()$FDR
contrProtRLMModerated$get_contrasts()$protein_Id |> unique() |> length()

```

```{r benchmarkModelRLM11Moderation}

ttd <- prolfqua::ionstar_bench_preprocess(contrProtRLMModerated$get_contrasts())
benchmark_ProtRLMModerated <- prolfqua::make_benchmark(
    ttd$data,
    model_description = "med. polish and rlm moderated",
    model_name = "prolfqua_rlm_mod")

prolfqua::table_facade(
  benchmark_ProtRLMModerated$smc$summary,
  caption = "Nr of proteins with Nr of not estimated contrasts.",
  digits = 1)

benchmark_ProtRLMModerated$plot_ROC(xlim = 1)
benchmark_ProtRLMModerated$plot_FDRvsFDP()

```
