---
title: "Benchmarking prolfqua across multiple datasets"
author: "Witold E Wolski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      toc_collapsed: true
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Benchmarking prolfqua across multiple datasets}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(prolfquabenchmark)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

create_mq_peptide_config <- function() {
  config <- prolfqua::AnalysisConfiguration$new()
  config$file_name <- "raw.file"
  config$hierarchy[["protein_Id"]] <- "leading.razor.protein"
  config$hierarchy[["peptide_Id"]] <- "sequence"
  config$hierarchy_depth <- 1
  config$ident_q_value <- "pep"
  config$set_response("peptide.intensity")
  config$isotope_label <- "isotope"
  config$min_peptides_protein <- 2
  config
}
```

# Setup

We benchmark the prolfqua `ContrastsLMMissingFacade` (LM + missing-value
merge + variance moderation) and `ContrastsLimmaFacade` (Limma empirical
Bayes) methods across four datasets.

```{r sharedSetup}
ionstar_ground_truth <- list(
  id_column = "protein_id",
  positive = list(label = "ECOLI", pattern = "ECOLI"),
  negative = list(label = "HUMAN", pattern = "HUMAN")
)

ionstar_contrasts <- c(
  "dilution_(9/7.5)_1.2"   = "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25"  = "dilution.d - dilution.c",
  "dilution_(6/4.5)_1.3(3)" = "dilution.c - dilution.b",
  "dilution_(4.5/3)_1.5"   = "dilution.b - dilution.a"
)

datadir <- file.path(find.package("prolfquadata"), "quantdata")
inputMQfile <- file.path(datadir, "MAXQuant_IonStar2018_PXD003881.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")
```

```{r helperFunction}
#' Run prolfqua benchmark with both Facade methods
#'
#' @param lfqdata_protein protein-level LFQData (log2-transformed, normalized)
#' @param formula_str model formula as character (e.g. "~ dilution.")
#' @param contrasts named character vector of contrasts
#' @param path_prefix directory prefix for output (e.g. "../inst/Benchresults/prolfqua_IonStar_MQ")
#' @param dataset_name dataset identifier for metadata
#' @param input_file input file description for metadata
#' @param ground_truth ground truth specification list
run_prolfqua_benchmark <- function(lfqdata_protein, formula_str, contrasts,
                                   path_prefix, dataset_name, input_file,
                                   ground_truth) {
  sw_version <- paste0("prolfqua ", packageVersion("prolfqua"))
  results <- list()

  # LM + missing + moderation
  fa_lm <- prolfqua::ContrastsLMMissingFacade$new(
    lfqdata_protein, formula_str, contrasts)
  results$lm <- fa_lm$get_contrasts()
  write_contrast_results(
    results$lm,
    path = paste0(path_prefix, "_lm"),
    metadata = list(
      dataset = dataset_name,
      method = "prolfqua_lm",
      method_description = "prolfqua LM + missing merge + moderation",
      input_file = input_file,
      software_version = sw_version,
      date = as.character(Sys.Date()),
      ground_truth = ground_truth))

  # Limma empirical Bayes
  fa_limma <- prolfqua::ContrastsLimmaFacade$new(
    lfqdata_protein, formula_str, contrasts)
  results$limma <- fa_limma$get_contrasts()
  write_contrast_results(
    results$limma,
    path = paste0(path_prefix, "_limma"),
    metadata = list(
      dataset = dataset_name,
      method = "prolfqua_limma",
      method_description = "prolfqua Limma empirical Bayes",
      input_file = input_file,
      software_version = sw_version,
      date = as.character(Sys.Date()),
      ground_truth = ground_truth))

  results
}
```


# Dataset 1: IonStar / MaxQuant peptides

```{r ds1LoadData}
mqdata <- list()
mqdata$data <- prolfquapp::tidyMQ_Peptides(inputMQfile)
mqdata$config <- create_mq_peptide_config()

annotation <- readxl::read_xlsx(inputAnnotation)
res <- dplyr::inner_join(mqdata$data, annotation, by = "raw.file")

mqdata$config$factors[["dilution."]] <- "sample"
mqdata$config$factors[["run_Id"]] <- "run_ID"
mqdata$config$factor_depth <- 1
mqdata$data <- prolfqua::setup_analysis(res, mqdata$config)

lfqdata_mq <- prolfqua::LFQData$new(mqdata$data, mqdata$config)
lfqdata_mq$set_data(lfqdata_mq$data_long() |> dplyr::filter(!grepl("^REV__|^CON__", protein_Id)))
lfqdata_mq$filter_proteins_by_peptide_count()
lfqdata_mq$remove_small_intensities()
```

```{r ds1Normalize}
tr <- lfqdata_mq$get_Transformer()
subset_h <- lfqdata_mq$get_copy()
subset_h$set_data(subset_h$data_long() |> dplyr::filter(grepl("HUMAN", protein_Id)))
subset_h <- subset_h$get_Transformer()$log2()$lfq
lfqdataNorm_mq <- tr$log2()$robscale_subset(lfqsubset = subset_h)$lfq
```

```{r ds1Aggregate}
agg_mq <- lfqdataNorm_mq$get_Aggregator()
agg_mq$aggregate()
lfqProt_mq <- agg_mq$lfq_agg
```

```{r ds1Fit}
results_mq <- run_prolfqua_benchmark(
  lfqdata_protein = lfqProt_mq,
  formula_str = "~ dilution.",
  contrasts = ionstar_contrasts,
  path_prefix = "../inst/Benchresults/prolfqua_IonStar_MQ",
  dataset_name = "IonStar_MQ",
  input_file = "MAXQuant_IonStar2018_PXD003881.zip/peptides.txt",
  ground_truth = ionstar_ground_truth)
```


# Dataset 2: IonStar / FragPipe MSstats.csv

```{r ds2LoadData}
inputFragfile <- file.path(datadir, "MSFragger_IonStar2018_PXD003881.zip")
annotation <- readxl::read_xlsx(inputAnnotation)

msstats <- read.csv(
  unz(inputFragfile, "IonstarWithMSFragger/MSstats.csv"),
  header = TRUE, sep = ",", stringsAsFactors = FALSE)
msstats <- tibble::tibble(msstats)
msstats$Run <- tolower(msstats$Run)

merged <- dplyr::inner_join(annotation, msstats, by = c("raw.file" = "Run"))
summedPep <- merged |>
  dplyr::group_by(raw.file, run_ID, sample, ProteinName, PeptideSequence) |>
  dplyr::summarize(nprecursor = dplyr::n(),
                   pepIntensity = sum(Intensity, na.rm = TRUE),
                   .groups = "drop")
```

```{r ds2Config}
atable <- prolfqua::AnalysisConfiguration$new()
atable$file_name <- "raw.file"
atable$hierarchy[["protein_Id"]] <- c("ProteinName")
atable$hierarchy[["peptide_Id"]] <- c("PeptideSequence")
atable$hierarchy_depth <- 1
atable$set_response("pepIntensity")
atable$factors[["dilution."]] <- "sample"
atable$factors[["run"]] <- "run_ID"
atable$factor_depth <- 1

config_fp <- atable
adata_fp <- prolfqua::setup_analysis(summedPep, config_fp)
lfqdata_fp <- prolfqua::LFQData$new(adata_fp, config_fp)
lfqdata_fp$filter_proteins_by_peptide_count()
lfqdata_fp$remove_small_intensities()
```

```{r ds2Normalize}
subset_h <- lfqdata_fp$get_copy()$get_Transformer()$log2()$lfq
subset_h$set_data(subset_h$data_long() |> dplyr::filter(grepl("HUMAN", protein_Id)))
tr <- lfqdata_fp$get_Transformer()
lfqdataNorm_fp <- tr$log2()$robscale_subset(lfqsubset = subset_h)$lfq

agg_fp <- lfqdataNorm_fp$get_Aggregator()
agg_fp$aggregate()
lfqProt_fp <- agg_fp$lfq_agg
```

```{r ds2Fit}
results_fp <- run_prolfqua_benchmark(
  lfqdata_protein = lfqProt_fp,
  formula_str = "~ dilution.",
  contrasts = ionstar_contrasts,
  path_prefix = "../inst/Benchresults/prolfqua_IonStar_FragPipe_MSstatsCSV",
  dataset_name = "IonStar_FragPipe_MSstatsCSV",
  input_file = "MSFragger_IonStar2018_PXD003881.zip/IonstarWithMSFragger/MSstats.csv",
  ground_truth = ionstar_ground_truth)
```


# Dataset 3: IonStar / FragPipe combined_protein.tsv

This dataset is already at protein level, so no peptide aggregation is needed.

```{r ds3LoadData}
annotation <- readxl::read_xlsx(inputAnnotation)
protein <- tibble::as_tibble(read.csv(
  unz(inputFragfile, "IonstarWithMSFragger/combined_protein.tsv"),
  header = TRUE, sep = "\t", stringsAsFactors = FALSE))

protein <- prolfquapp::tidy_FragPipe_combined_protein_deprec(protein)
protein <- protein |> dplyr::filter(unique.stripped.peptides > 1)
merged <- dplyr::inner_join(annotation, protein)
```

```{r ds3Config}
atable3 <- prolfqua::AnalysisConfiguration$new()
atable3$file_name <- "raw.file"
atable3$hierarchy[["protein_Id"]] <- c("protein")
atable3$hierarchy_depth <- 1
atable3$set_response("razor.intensity")
atable3$factors[["dilution."]] <- "sample"
atable3$factors[["run"]] <- "run_ID"
atable3$factor_depth <- 1

config3 <- atable3
adata3 <- prolfqua::setup_analysis(merged, config3)
lfqdata3 <- prolfqua::LFQData$new(adata3, config3)
lfqdata3$remove_small_intensities()
```

```{r ds3Normalize}
subset_h <- lfqdata3$get_copy()$get_Transformer()$log2()$lfq
subset_h$set_data(subset_h$data_long() |> dplyr::filter(grepl("HUMAN", protein_Id)))
tr <- lfqdata3$get_Transformer()
lfqNorm3 <- tr$log2()$robscale_subset(lfqsubset = subset_h)$lfq
```

```{r ds3Fit}
results_cp <- run_prolfqua_benchmark(
  lfqdata_protein = lfqNorm3,
  formula_str = "~ dilution.",
  contrasts = ionstar_contrasts,
  path_prefix = "../inst/Benchresults/prolfqua_IonStar_FragPipe_combined_protein",
  dataset_name = "IonStar_FragPipe_combined_protein",
  input_file = "MSFragger_IonStar2018_PXD003881.zip/IonstarWithMSFragger/combined_protein.tsv",
  ground_truth = ionstar_ground_truth)
```


# Dataset 4: CPTAC

```{r ds4Setup}
evalCPTAC <- requireNamespace("msdata", quietly = TRUE)
```

```{r ds4Notice, echo=FALSE, results='asis'}
if (!evalCPTAC) {
  cat("> **Note:** The CPTAC section requires the `msdata` package, which is not currently installed.",
      "CPTAC code chunks are skipped. Install it and re-render to see results.\n")
}
```

```{r ds4LoadData, eval = evalCPTAC}
tmp <- msdata::quant(full.names = TRUE)
xx <- read.csv(tmp, sep = "\t")
peptides <- prolfquapp::tidyMQ_Peptides(xx)

annotation_cptac <- data.frame(raw.file = unique(peptides$raw.file))
annotation_cptac <- annotation_cptac |>
  dplyr::mutate(group = gsub("^6", "", raw.file)) |>
  dplyr::mutate(group = gsub("_[0-9]$", "", group))

peptides <- dplyr::inner_join(annotation_cptac, peptides)

config_cptac <- create_mq_peptide_config()
config_cptac$factors["group."] <- "group"

peptides <- prolfqua::setup_analysis(peptides, config_cptac)
lfqpeptides <- prolfqua::LFQData$new(peptides, config_cptac)
lfqpeptides$filter_proteins_by_peptide_count()
lfqpeptides$remove_small_intensities()
```

```{r ds4Transform, eval = evalCPTAC}
tr <- lfqpeptides$get_Transformer()
tr <- tr$intensity_matrix(vsn::justvsn)
lfqtransformed <- tr$lfq
agg <- lfqtransformed$get_Aggregator()
agg$aggregate()
lfqProt_cptac <- agg$lfq_agg
```

```{r ds4Fit, eval = evalCPTAC}
cptac_contrasts <- c("b_vs_a" = "group.b - group.a")
cptac_ground_truth <- list(
  id_column = "protein_id",
  positive = list(label = "UPS", pattern = "_UPS"),
  negative = list(label = "YEAST", pattern = "_YEAST")
)

results_cptac <- run_prolfqua_benchmark(
  lfqdata_protein = lfqProt_cptac,
  formula_str = "~ group.",
  contrasts = cptac_contrasts,
  path_prefix = "../inst/Benchresults/prolfqua_CPTAC",
  dataset_name = "CPTAC",
  input_file = "msdata::quant()/peptides.txt",
  ground_truth = cptac_ground_truth)
```
