---
title: "Comparing squeezeVar: limma, msqrob2, and prolfqua"
author: "Witold E Wolski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      toc_collapsed: true
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Comparing squeezeVar: limma, msqrob2, and prolfqua}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4)
run <- requireNamespace("limma", quietly = TRUE)
```

## Why this vignette exists

`prolfqua` used to ship its own empirical-Bayes variance-moderation routine, `squeezeVarRob()`.
The source header records where it came from:

> `# Code from https://raw.githubusercontent.com/statOmics/MSqRob/master/R/squeezeVarRob.R`

That is the **old** MSqRob package (the 2016-era predecessor of `msqrob2`). It was never entirely
clear *why* MSqRob carried its own copy of a function that already exists in `limma`. This vignette
answers that empirically by running the three implementations side by side:

1. `limma::squeezeVar()` — the reference implementation.
2. `msqrob2` — the current Bioconductor package.
3. `prolfqua::squeezeVarRob()` — the former prolfqua backend, inherited from old MSqRob.

> **Outcome.** The benchmark in this vignette (synthetic + real IonStar spike-in data) shows the two
> estimators produce near-identical results, so **`prolfqua` has migrated its variance moderation to
> `limma::squeezeVar()`** (`moderated_p_limma()` as of prolfqua 1.6.3), and the vendored
> `squeezeVarRob()` has been **removed**. This vignette reconstructs its behaviour from limma
> primitives (`limma::squeezeVar(legacy = TRUE)` + a `min_df` filter) for the comparison below.

The short answer, established below:

- **`msqrob2` does not have its own implementation.** It calls `limma::squeezeVar()` directly, so
  it *is* limma.
- **The former `prolfqua::squeezeVarRob()` was limma's classic ("legacy") `fitFDist` plus one
  addition: a `min_df` pre-filter (default `1`)** that drops features whose residual degrees of freedom
  are `<= min_df` before estimating the prior.
- Old MSqRob added that filter because the classic estimator **mis-estimates the prior (collapses to
  `df.prior = Inf`, i.e. no moderation) when the data contain many low / fractional-df features** —
  exactly what peptide-level robust/ridge regression produces. **Modern limma fixed the same problem
  upstream** with a redesigned estimator (the `legacy` argument), which is why `msqrob2` needs no
  patch.
- **This regime is live in `prolfqua`, not hypothetical.** `prolfqua` feeds its variance moderation the
  per-contrast residual df, and several of its estimators produce **fractional or low df**: `rlm` and
  `rfit` (df = sum of robust weights − rank), `lmer_nested` (Satterthwaite df on peptide-level mixed
  models), and the imputation facades (`lm_impute` / `lm_missing`, with borrowed / low df). Only plain
  `lm` gives integer df. So the behavioural difference between the classic-plus-`min_df` fork and
  modern limma **can change prolfqua results** on exactly those paths — it is not merely historical.
  (`firth` and `firth_nested` are *not* consumers — they use their own `ContrastsFirth` inference, not
  `squeezeVarRob`.)

```{r libs}
library(limma)
library(prolfqua)
library(ggplot2)
```

## msqrob2 delegates to limma

`msqrob2::msqrobLm()` and `msqrob2::msqrobLmer()` compute the moderated variance with a plain call to
`limma::squeezeVar()` (default arguments: non-robust, no covariate):

```{r msqrob2_source, eval = requireNamespace("msqrob2", quietly = TRUE)}
body_lines <- deparse(getNamespace("msqrob2")$msqrobLm)
cat(body_lines[grep("squeezeVar", body_lines)[1] + 0:1], sep = "\n")
```

So "the msqrob2 implementation" is `limma::squeezeVar(var, df)`. We keep it as a separate column
below only to make the three-way comparison explicit; it is identical to the limma column by
construction.

## The three implementations

The former `prolfqua::squeezeVarRob()` was **removed** in prolfqua 1.6.3, so we reproduce it here from
limma primitives — which is exactly what it was: limma's **classic** estimator (`legacy = TRUE`) plus
MSqRob's `min_df` pre-filter (drop features with residual df ≤ `min_df` before estimating the prior,
then apply the fitted prior to all features).

```{r impls}
recycle <- function(x, k) if (length(x) == 1L) rep(x, k) else x

# Exact reconstruction of the removed prolfqua::squeezeVarRob() (non-robust path).
former_fork <- function(var, df, min_df = 1) {
  n <- length(var)
  dfv <- recycle(df, n)
  keep <- is.finite(dfv) & dfv > min_df
  fit <- limma::squeezeVar(var[keep], df = dfv[keep], legacy = TRUE)  # limma classic fitFDist
  var.post <- if (is.infinite(fit$df.prior)) {
    rep(fit$var.prior, n)                                             # full collapse -> prior variance
  } else {
    (dfv * var + fit$df.prior * fit$var.prior) / (dfv + fit$df.prior)
  }
  list(df.prior = fit$df.prior, var.prior = fit$var.prior, var.post = var.post)
}

run_three <- function(var, df) {
  n <- length(var)
  lim  <- limma::squeezeVar(var, df)          # modern default -> what prolfqua now uses
  msq  <- limma::squeezeVar(var = var, df = df)  # exactly the msqrob2 call
  pro  <- former_fork(var, df)                 # the removed squeezeVarRob fork
  tibble::tibble(
    df               = df,
    var              = var,
    varpost_limma    = lim$var.post,
    varpost_msqrob2  = msq$var.post,
    varpost_prolfqua = pro$var.post,
    dfprior_limma    = recycle(lim$df.prior, n),
    dfprior_prolfqua = recycle(pro$df.prior, n),
    varprior_limma   = recycle(lim$var.prior, n),
    varprior_prolfqua = recycle(pro$var.prior, n)
  )
}

# scaled-F generative model that the estimators assume:
# per-feature true variance sigma^2 ~ scaled inverse chi-square(df_prior, var_prior),
# observed sample variance ~ sigma^2 * chisq(df) / df.
gen_scaled_F <- function(n, df, df_prior = 4, var_prior = 0.4) {
  sigma2 <- var_prior * df_prior / rchisq(n, df_prior)
  sigma2 * rchisq(n, df) / df
}

summarise_cmp <- function(d) {
  data.frame(
    dfprior_limma     = d$dfprior_limma[1],
    dfprior_prolfqua  = d$dfprior_prolfqua[1],
    varprior_limma    = d$varprior_limma[1],
    varprior_prolfqua = d$varprior_prolfqua[1],
    max_abs_varpost_diff  = max(abs(d$varpost_limma - d$varpost_prolfqua)),
    mean_rel_varpost_diff = mean(abs(d$varpost_limma - d$varpost_prolfqua) / d$varpost_limma),
    msqrob2_equals_limma  = isTRUE(all.equal(d$varpost_limma, d$varpost_msqrob2))
  )
}
```

## Regime 1 — integer residual df (baseline)

The controlled baseline: 4000 features with integer df in `3:9`, as produced by plain `lm` fits. This
establishes what the estimators do when they agree.

```{r integer_df}
set.seed(7)
n  <- 4000
df_int <- sample(3:9, n, replace = TRUE)
var_int <- gen_scaled_F(n, df_int)

cmp_int <- run_three(var_int, df_int)

knitr::kable(summarise_cmp(cmp_int), digits = 6,
  caption = "Integer df: modern limma vs the former fork (msqrob2 == limma).")
```

The posterior variances are indistinguishable (mean relative difference `< 1e-3`). The former fork
matches limma's **legacy** algorithm exactly here (all df > 1, so the `min_df` filter is inert):

```{r integer_legacy}
lg <- limma::squeezeVar(var_int, df_int, legacy = TRUE)
pr <- former_fork(var_int, df_int)
c(dfprior_limma_legacy = lg$df.prior,
  dfprior_former_fork  = pr$df.prior,
  varprior_limma_legacy = lg$var.prior,
  varprior_former_fork  = pr$var.prior)
```

```{r integer_scatter, fig.cap = "Posterior variances agree on the diagonal (integer df, non-robust)."}
ggplot(cmp_int, aes(varpost_limma, varpost_prolfqua)) +
  geom_point(alpha = 0.2, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, colour = "red") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "var.post (limma)", y = "var.post (prolfqua)") +
  theme_bw()
```

## Regime 2 — fractional / low residual df (why MSqRob forked it — and why it matters here)

Robust and mixed-model fits produce **fractional** residual df, and sparse fits produce **low** df.
This is the regime where the classic estimator misbehaves. We simulate df drawn uniformly from
`[0.3, 4]`.

```{r frac_df}
set.seed(7)
df_frac  <- runif(n, 0.3, 4)
var_frac <- gen_scaled_F(n, pmax(df_frac, 0.3))
cat("features with df <= 1:", sum(df_frac <= 1), "of", n, "\n")

cmp_frac <- run_three(var_frac, df_frac)
knitr::kable(summarise_cmp(cmp_frac), digits = 6,
             caption = "Fractional df: limma (modern default) vs prolfqua.")
```

Here the two diverge. The reason is not merely the `min_df` threshold — it is that the former fork was
limma's **classic** estimator, while modern `limma::squeezeVar()` defaults to a **redesigned** one. We
can expose all the moving parts:

```{r frac_decompose}
show <- function(tag, x) {
  data.frame(source = tag, df.prior = x$df.prior[1], var.prior = x$var.prior[1])
}
rbind(
  show("limma default (new estimator)",      limma::squeezeVar(var_frac, df_frac)),
  show("limma legacy = TRUE (classic)",       limma::squeezeVar(var_frac, df_frac, legacy = TRUE)),
  show("former fork, min_df = 1e-15 (none)",  former_fork(var_frac, df_frac, min_df = 1e-15)),
  show("former fork, min_df = 1 (default)",   former_fork(var_frac, df_frac))
)
```

Two exact identities fall out of this table:

- **former fork with `min_df = 1e-15` == `limma(legacy = TRUE)`** — both return `df.prior = Inf`.
  This confirms the former `squeezeVarRob()` *was* limma's classic `fitFDist`. On this fractional data
  the classic estimator **collapses to `df.prior = Inf`**: no shrinkage, every feature falls back to a
  pooled variance. That is the failure mode MSqRob hit.
- **former fork with `min_df = 1`** (the shipped default) recovers a sensible finite `df.prior` by
  excluding the `<= 1`-df features from the prior fit. That is precisely the patch MSqRob added.
- **Modern `limma` (default)** also recovers a sensible `df.prior`, but via its new algorithm rather
  than a df filter.

```{r frac_dfprior_bar, fig.cap = "Estimated prior df on fractional-df data: classic limma collapses to Inf; the min_df patch and modern limma both recover a finite value."}
dfp <- data.frame(
  method = factor(
    c("limma\n(new default)", "limma\n(legacy)", "former fork\n(min_df=1e-15)", "former fork\n(min_df=1)"),
    levels = c("limma\n(legacy)", "former fork\n(min_df=1e-15)", "limma\n(new default)", "former fork\n(min_df=1)")),
  df.prior = c(limma::squeezeVar(var_frac, df_frac)$df.prior,
               limma::squeezeVar(var_frac, df_frac, legacy = TRUE)$df.prior,
               former_fork(var_frac, df_frac, min_df = 1e-15)$df.prior,
               former_fork(var_frac, df_frac)$df.prior)
)
dfp$label <- ifelse(is.finite(dfp$df.prior), sprintf("%.2f", dfp$df.prior), "Inf")
dfp$plot_val <- ifelse(is.finite(dfp$df.prior), dfp$df.prior, max(dfp$df.prior[is.finite(dfp$df.prior)]) * 1.3)
ggplot(dfp, aes(method, plot_val, fill = is.finite(df.prior))) +
  geom_col() +
  geom_text(aes(label = label), vjust = -0.3) +
  scale_fill_manual(values = c(`TRUE` = "steelblue", `FALSE` = "firebrick"),
                    labels = c(`TRUE` = "finite", `FALSE` = "Inf (collapsed)"), name = NULL) +
  labs(x = NULL, y = "estimated prior df") +
  theme_bw()
```

## Where prolfqua feeds the squeeze

There is a single call site in `prolfqua`: `moderated_p_limma()` (in `tidyMS_moderation.R`), which
squeezes the per-contrast residual variances. It now calls `limma::squeezeVar()` (the migration this
vignette justifies); it previously called the vendored `squeezeVarRob()`:

```r
# current (prolfqua >= 1.6.3)
squeezed_var <- limma::squeezeVar(contrast_df$sigma^2, df = contrast_df[[df]], robust = robust)
# former
# squeezed_var <- prolfqua::squeezeVarRob(contrast_df$sigma^2, df = contrast_df[[df]], robust = robust)

# prior degrees of freedom are Inf
if (all(is.infinite(squeezed_var$df.prior))) {
  squeezed_var$df.prior <- mean(contrast_df[[df]]) * nrow(contrast_df) / 10
}
```

The fallback on the second line is itself evidence: `prolfqua` already has to catch the
`df.prior = Inf` **collapse** on real data. `moderated_p_limma()` is called by `ContrastsModerated`
(the moderated-Wald path wrapping `lm`, `rlm`, `rfit`, the `lm_impute` / `lm_missing` / `rfit_impute`
facades, and `lmer_nested`) and by the DEqMS small-group fallback.

The `df` column fed in is the per-contrast residual df, and it is **not** integer for several
estimators:

- **`rlm`, `rfit`** — the robust residual df is `sum(robust weights) − rank`; the weights lie in
  `(0, 1]`, so the df is **fractional**.
- **`lmer_nested`** — the contrast df is the **Satterthwaite** denominator df from
  `lmerTest::contest()`, **fractional** and, on peptide-level nested models, often small.
- **`lm_impute` / `lm_missing` / `rfit_impute`** — residual df can be **low or borrowed** when groups
  are (partly) imputed, where the `min_df = 1` filter changes which features enter the prior fit.
- **`lm`** — integer residual df; the only path that stays in Regime 1.

(`firth` / `firth_nested` do **not** feed `squeezeVarRob` — they use their own `ContrastsFirth`
inference. Likewise `limma*` uses limma's own `eBayes`, and `limpa` its own moderation.)

A quick check that `rlm` really produces fractional df:

```{r rlm_frac_df}
set.seed(1)
x <- rnorm(12); y <- 2 * x + rnorm(12); y[3] <- y[3] + 8   # inject one outlier
m <- MASS::rlm(cbind(1, x), y)
c(rank = m$rank, sum_weights = sum(m$w), rlm_df_residual = sum(m$w) - m$rank)
```

So the classic-vs-modern difference shown above is reachable in ordinary `prolfqua` usage —
`rlm` / `rfit` and `lmer_nested` on peptide-level data, and the imputation facades — not only in a
synthetic peptide-ridge scenario.

## So why did MSqRob roll its own?

Putting the pieces together:

- Around 2016, MSqRob copied limma's `fitFDist` / `squeezeVar` and added a **`min_df` guard**.
- The motivation: MSqRob fits **peptide-level robust ridge regressions**, whose residual degrees of
  freedom are fractional and frequently `< 1`. On such data, limma's estimator of that era collapsed
  to `df.prior = Inf` — effectively switching moderation off and returning pooled variances. Dropping
  the `<= min_df` features before the prior fit restores a usable, finite prior df.
- **Modern limma made the fork obsolete for limma's own users.** Recent limma (the `legacy`
  argument; default `legacy = NULL` selects the new estimator for non-integer df) handles fractional
  df robustly on its own. `msqrob2` therefore dropped the custom code and simply calls
  `limma::squeezeVar()`.
- **For `prolfqua` the fork is still doing work.** `prolfqua` inherited the classic estimator plus
  the `min_df` patch and applies it to `rlm` / `rfit` (fractional df), `lmer_nested` (fractional
  Satterthwaite df), and the imputation facades (low / borrowed df) — precisely the regime where
  classic limma collapses. The `min_df` filter is what keeps those paths from falling back to
  unmoderated pooled variances (with a further explicit `Inf`-collapse guard in `moderated_p_limma()`).
  It only leaves `lm` unaffected.

## Does it matter? Benchmark

The algorithm analysis above shows the fork and modern limma *diverge* on fractional-df paths. Whether
that divergence *matters* for real conclusions is an empirical question. Both backends were run through
the **same** models, contrasts and downstream statistics — only the squeeze backend varied — on two
data sources:

- **Synthetic** — 30 replicates × 2 noise scenarios (~300 proteins each) with known regulated proteins.
- **Real IonStar** spike-in data (`prolfquadata`, PXD003881): *E. coli* (true positives) spiked into a
  constant human background (true negatives); 4178 proteins, 4 dilution contrasts.

Three moderated facades were compared, spanning the df regimes: `lm` (integer df), `rlm` (robust,
fractional df) and `lmer_nested` (peptide-level mixed model, fractional Satterthwaite df). Results are
cached in the package (fitting IonStar `lmer_nested` on all proteins is too slow to run at build time).

```{r load_bench}
bench_file <- system.file("extdata", "squeezeVar_benchmark.rds", package = "prolfquabenchmark")
if (!nzchar(bench_file)) {
  # dev / build-time fallback: vignettes/ sits next to inst/
  alt <- file.path("..", "inst", "extdata", "squeezeVar_benchmark.rds")
  if (file.exists(alt)) bench_file <- alt
}
bench <- if (nzchar(bench_file) && file.exists(bench_file)) readRDS(bench_file) else NULL
```

### The divergence regime is genuinely reached

On real IonStar data, `rlm` and `lmer_nested` feed overwhelmingly fractional df into the squeeze,
while `lm` stays integer — so any classic-vs-modern difference is exercised, not hypothetical:

```{r df_regime}
if (!is.null(bench)) knitr::kable(bench$ionstar$df_regime, caption = "IonStar: residual df fed into the squeeze, by method.")
```

### Real IonStar: ranking, calibration, per-protein calls

```{r ionstar_metrics}
if (!is.null(bench)) knitr::kable(bench$ionstar$metrics,
  caption = "IonStar @ BH q <= 0.05. AUC = ranking; FDP_at_0.05 = observed false-discovery proportion (E.coli/HUMAN truth); power = fraction of E.coli called.")
```

```{r ionstar_concordance}
if (!is.null(bench)) knitr::kable(bench$ionstar$concordance,
  caption = "IonStar: agreement of significance calls (q <= 0.05) between the two backends.")
```

AUC is identical to three decimals, the observed false-discovery proportion tracks the same value for
both backends, and **≥ 99.7% of significance calls are identical** — only 5 (`lm`), 30 (`rlm`) and 43
(`lmer_nested`) of ~16,000 protein×contrast calls flip when switching backends.

### Synthetic replicates

```{r synthetic_metrics}
if (!is.null(bench)) knitr::kable(bench$synthetic$metrics, digits = 3,
  caption = "Synthetic, mean over 30 replicates @ BH q <= 0.05.")
```

The synthetic replicates agree with IonStar: AUC, power and FDP are near-identical between backends in
every cell. (The synthetic `lmer_nested` FDP is inflated for *both* backends because `sim_lfq_data`'s
peptide mode redraws peptide baselines per group — an anti-conservative simulator artifact that cancels
in the backend-vs-backend comparison; the real IonStar FDP above is the calibration-trustworthy number.)

## Conclusion

| Implementation | What it is | Status |
|---|---|---|
| `limma::squeezeVar()` (default) | modern reference estimator; handles fractional df natively | **what prolfqua now uses** |
| `msqrob2` | calls `limma::squeezeVar(var, df)` | identical to limma |
| `prolfqua::squeezeVarRob()` (former) | limma **classic** `fitFDist` + `min_df` filter (from MSqRob) | **removed** from prolfqua |

The fork and modern limma agree on the integer-df (`lm`) path and *diverge algorithmically* on the
fractional/low-df paths (`rlm` / `rfit`, `lmer_nested`, imputation facades) — but the benchmark shows
that divergence is **immaterial to conclusions**: identical AUC, maintained FDR calibration, and
≥ 99.7% identical calls on gold-standard spike-in data.

Because modern `limma::squeezeVar()` solves the same fractional-df collapse without the `min_df`
heuristic — and is actively maintained, whereas the fork was a static 2016 copy — **prolfqua migrated
its variance moderation to `limma::squeezeVar()`** (`moderated_p_limma()`, prolfqua 1.6.3), and
the vendored `squeezeVarRob()` has been removed. This vignette is the evidence behind that decision.

```{r session}
sessionInfo()
```
